---
#title: The Viridian Project
#subtitle: For a Sustainable and Fair Economy
author: mvoge
date: '2019-03-24'
---

## For a Sustainable and Fair Economy

We live in a world where many things seem to be **going wrong**. Growing
social inequality, climate change, loss of biodiversity... just to name a few.
**Are you one of those people thinking that change is needed, however nothing ever happens?**
Then maybe this project is for you.

Most of the problems of human society have to do with economics.
This is why the *Viridian Project* aims for an alternative way of **economic activity**.
Imagine an economy that takes into account not only economical, but also
environmental and social concerns. An economy where people are worth more than
profits. An economy, in which **sustainable and fair are the default**, not the
exception.

<!--
The inspiration for the Viridian Project comes from two communities having little
contact with each other so far: The vision of the *Transition Town* movement of
a **sustainable, post-fossil, resilient economy** on one side. The vision of the
*cryptocurrency community* of a **decentralized peer-to-peer economy** on the other
side.
-->

The main goal of the Viridian Project is to **internalize external costs using a new digital currency**,
making social and environmental concerns a normal part of every business model.
<!--
This incentivizes businesses that are not harmful for
society. When social peer pressure is included into economic activity (via the
alternative digital currency), people will act more responsibly automatically.
This is similar to the way that *commons* work, where a group of people
collectively cares for a common resource, making every individual act responsibly.
-->

<!--
To achieve this, a computer network shall be implemented, in which the desired
incentives are algorithmically inherent and governed collectively by the crowd.
These algorithms mut be very carefully designed, transparent and
agreed upon by everyone. They must also be continually debated and adjusted
as soon as needed.
-->

<a href="../paper" class="btn btn-info" role="button">Read the whitepaper</a>

## Goals

The objective is to create a computer network for economic transactions providing
the following:

<div markdown class="row">
<div markdown class="col-md-6">

![Decentralization](/images/home/network.svg){style="height: 75px;"}

The network shall be decentralized for increased **resilience**.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-dec" aria-expanded="false" aria-controls="showmore-dec"><i class="fa fa-caret-down" aria-hidden="true"></i> Show more</button>

<!-- <i class="glyphicon glyphicon-triangle-bottom"></i> -->
<!-- &#x25bc; UTF-8 alternative to glyphicon-triangle-bottom -->

<p></p>

<div class="collapse" id="showmore-dec">

There should not be any single points of failure. It must also be spread to
**prevent too much concentration of power**. The network shall be operated by
and in the hands of all people, not of few (or single) corporations or governments.

</div>

<!-- <div>Icon made by me!</div> -->

</div>
<div class="col-md-6">

<div class="figure">
![Democratization](/images/home/like.svg){style="height: 75px;"}
![Democratization](/images/home/dislike.svg){style="height: 75px;"}
<p class="caption">Democratization</p>
</div>

Thanks to decentralization economy and society become
**more democratic and more equal**.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-dem" aria-expanded="false" aria-controls="showmore-dem"><i class="fa fa-caret-down" aria-hidden="true"></i> Show more</button>

<p></p>

<div class="collapse" id="showmore-dem">

Running the network on many decentralized nodes, the concentration of resources
and power can be prevented. Decentralized systems like e.g. blockchains
enable to **move power** away from centralized systems (like banks and corporations)
and towards the people using the infrastructure. In addition, the
rules of the network, more specifically the assessment of products and companies,
shall be **set by the users**: in an open, collaborative process, similar to the
development of open source software. The use of computer networks allows the usage
of innovative democratic consensus tools like liquid democracy.

</div>

<!-- <div>Icons made by <a href="https://www.flaticon.com/authors/smashicons" title="Smashicons">Smashicons</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
</div>

####

<div class="row">
<div class="col-md-6">

![Sustainability](/images/home/sprout.svg){style="height: 75px;"}

Transparent pricing mechanisms **incentivize** sustainable and socially responsible business.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-sust" aria-expanded="false" aria-controls="showmore-sust"><i class="fa fa-caret-down" aria-hidden="true"></i> Show more</button>

<p></p>

<div class="collapse" id="showmore-sust">

Sustainable and socially responsible business and commerce are incentivized
**by including external costs** of economic activity in the prices. If a
company generates high external costs that is does not take care to prevent or
mitigate, it often receives an **unfair competitive advantage**. But the costs still
arise, for us all (often the taxpayer). They are socialized. If prices include
external costs from the outset, companies gain a competitive advantage if they
reduce them instead of socializing costs more and more.
Socially meaningful solutions are automatically preferred.

In addition, for sustainability reasons, it is important that the blockchain
transactions are not verified in a resource-intensive proof-of-work process,
but in an alternative way with an ecological footprint as small as possible.

</div>

<!-- <div>Icons made by <a href="https://www.flaticon.com/authors/retinaicons" title="Retinaicons">Retinaicons</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
<div class="col-md-6">

![Fairness](/images/home/handshake.svg){style="height: 75px;"}

The Viridian market will be fairer and **closer to an idealized free market**.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-fair" aria-expanded="false" aria-controls="showmore-fair"><i class="fa fa-caret-down" aria-hidden="true"></i> Show more</button>

<p></p>

<div class="collapse" id="showmore-fair">

If external costs are included in the price calculation,
the market is fairer and is approaching an idealized free market.
On such a market, **all market participants are perfectly informed**.
The resulting prices reflect the **costs incurred to society** through production.
In other words, everyone affected by the production will receive adequate
compensation. Because if someone is not compensated appropriately, they can simply
go elsewhere on a free market. In case of external costs that does not work.
The market is imperfect and Viridian closes this gap a bit.

Also, because the supply chain can be revealed in the process, inferior quality
can be revealed as well. Customers do not have to trust the reputation of a
producer, but can choose the best option available. This makes it easier for
newcomers to sell their better products. Especially if external costs are reduced.
This **accelerates the introduction of alternative technologies**, thereby increasing
innovation and reducing market monopolies.

</div>

<!-- <div>Icons made by <a href="https://www.flaticon.com/authors/zlatko-najdenovski" title="Zlatko Najdenovski">Zlatko Najdenovski</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
</div>

####

<div class="row">
<div class="col-md-6">

![Transparency](/images/home/search.svg){style="height: 75px;"}

It shall be **visible at all times how the rules of the network are**
and *why* they are like that.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-trans" aria-expanded="false" aria-controls="showmore-trans"><i class="fa fa-caret-down" aria-hidden="true"></i> Show more</button>

<p></p>

<div class="collapse" id="showmore-trans">

The mechanics of the network must be transparent at all times, including the
internalization of external costs. Changes to the algorithms must be
accompanied by legitimizing information.
**Public data shall be made open to everyone.** Because of its importance for
society, information on the external costs of any economic activity must be
public and cannot be kept a company secret. Because people are
more important than profits, there cannot be such "investor protection".

</div>

<!-- <div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
<div class="col-md-6">

![Privacy](/images/home/lock.svg){style="height: 75px;"}

The network shall be **end-to-end encrypted** to provide security
and privacy.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-priv" aria-expanded="false" aria-controls="showmore-priv"><i class="fa fa-caret-down" aria-hidden="true"></i> Show more</button>

<p></p>

<div class="collapse" id="showmore-priv">

People should have full control over their data.
**Private data shall be kept private**, instead of being a commodity,
which is harvested for profit.

</div>

<!-- <div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
</div>

<!-- * **Transactions:** Economic transactions shall be performed via the network, -->
<!--   i.e. tokens of value (a currency) shall be exchanged via the network. -->


## Get involved

There are many open questions and unsolved problems. At the moment, the project
is just an idea by a single person (see <a href="about.html">about</a> and
<a href="contact.html">contact</a>).
**I am looking for like-minded people to discuss and exchange ideas.**
Once the vision is shaped into a concrete network architecture and
algorithms/rules, one can start to work on implementations. But for the moment,
I would like to get feedback from as many people as possible, especially on the
subjects listed under the tiles below. 

**Are you inspired by the idea of the project? You can help!**
Just click on your area of interest and start getting involved.

```{r, echo=FALSE}
htmltools::includeHTML("snippets/tiles.html")
```

<div style="margin-bottom: 30px;"></div>

<div id="cryptocurrencies" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Cryptocurrencies</h3>
</div>
<div class="panel-body tile-panel-body">

### Which solution to choose?

Bitcoin, Litecoin, Ethereum, IOTA, blockchain, DAG, smart contracts, DAOs,
proof-of-work, proof-of-stake, proof-of-cooperation, proof-of-interest, ... ---
the world of cryptocurrencies is full of innovative ideas, with new ones
springing up monthly.

**If you are enthusiastic about all things crypto, then you might be able to help us decide, which solution is best suitable for the Viridian project.**

We want transaction validation to have an as small resource footprint as
possible. The incentive system must be designed such that there is no motivation
for running a lot of computing power, however there must be motivation to
contribute to the network in the long term. Of course, the system shall be as
decentralized and trustless as possible to keep it secure and scalable.

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="software-development" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Software development</h3>
</div>
<div class="panel-body tile-panel-body">

### Good design is the key

The Viridian project's concept relies heavily on the use of computer technology
and software. All this software still has to be built, both on the backend side
and on the frontend/UI/UX side.

**A clear vision and good design choices are crucial for an infrastructure that is secure and maintainable and that scales well.**

One should build upon existing work wherever possible. Especially when it comes
to cryptography, there is a good rule of thumb that one should not "come up
with a home-brew solution". Therefore, a first step would be, together with
creating the vision, to carefully evaluate which components could be based on
what, and how these components must be able to interact with each other, keeping
in mind maintainability, security and scalability.

Once these general points are sorted out and the roadmap is clear, one can set
up the requirements for the various components and can start implementing them
(perhaps just making adjustments or additions to already existing code).

A slick UI/UX will also be a crucial part of attracting users.

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="monetary" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Currencies and monetary policy</h3>
</div>
<div class="panel-body tile-panel-body">

### What are the rules of a good currency?

In traditional fiat currencies, a central bank is in charge of controlling the
currency, e.g. by regulating the amount of money in circulation, or in fact by
setting the key interest rate.

In a decentralized cryptocurrency, there is no central bank, which is a good thing
because it prevents corruption and dependency on centralized entities.
Instead, the network's mathematical algorithms take over the role of a central bank.

An important question is: **How to design these algorithmic rules?** For instance,
what is the **optimal supply of fresh money**? Shall there be a fixed amount of money?
Or shall the amount of money increase (causing inflation) or decrease? Shall
there be a demurrage imposed on the currency, constantly devaluating it over
time?

### What role will the financial system play?

In the Viridian economy, will banks still be important players? Private banks
currently create almost all the money that enters the economy. With a decentralized
cryptocurrency as dominant payment medium, this would not be the case anymore.

So, how shall banks be treated in the Viridian economy? Will banks' services also
have associated colors? **Will banks be superseded** by innovative new concepts, e.g.
peer-to-peer credit systems with risk scattering by spreading investment money
on many agents?

What to do with interest rates on investments? **Shall there still be interest rates?**
Or are interest rates not acceptable anymore since they are drivers of limitless
growth in a world with limited resources?

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="economics" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Economics</h3>
</div>
<div class="panel-body tile-panel-body">

### Would the Viridian economy be closer to an "ideal" free market?

The concept of *free market economy* goes back to the influential economists
Adam Smith and David Ricardo. Smith introduced the idea that a free market
results in fair conditions for all participants. Ricardo formalized and abstracted
Smith's ideas and had to simplify it a lot. This resulted in a well working
theory, but many people also started to neglect the limitations of the model
and the fact that it does not necessarily represent reality.

Adam Smith was very aware of the limitations of his concept and warned about them.
For instance, a market with few corporations is a bad idea because the corporations
can choose to require higher prices than they need to. It becomes even worse when
big business and government make each other gifts, abusing their power, leading
to high profits for few capitalist leaders and corrupt politics. This is why
Smith was a fierce opponent of corporations and preferred an economy with many
small competitors (like most of economy still was at Smith's time, which changed
over the course of the 19th century).

The Viridian economy, with its decentralized power and easily adjustable rules,
**might help to prevent such situations and create a market that is closer to Smith's ideal**.

Another flaw of our present-day capitalist economy is the exclusion of external
costs. Many economists argue that this leads to a distortion of the free market.
**In Smith's ideal free market, all prices reflect the actual cost that a product's production imposes on society.**
For example, if a worker needs to spend more time on one product than on another,
then it must cost more to reflect this. If the production leads to external costs
that society (ultimately everyone, including the buyer and the seller) must pay for,
then the product should also cost more, to pay for these costs and also to regulate
the demand for the product.

Again, the Viridian project might help to remove this flaw from the economy.

What do you think about it? If you have any opposing views or ideas how the
Viridian project can improve our market economy even more, feel free to contact
us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

### What are the likely conerns and needs of businesses?

The Viridian project would necessarily interfere with the way people are doing
business. For example, color must be considered when determining retail prices.

From the business perspective, what would the main **concerns and objections** be
that might disturb the adoption of Viridian? How could these concerns be addressed?
What are the **possibilities to make a transition easier for businesses**? Can we
provide services that would simplify a business's life?

### How to attract businesses?

How can we create business models that would
**make the adoption of Viridian lucrative for businesses**,
e.g. through improved image or intensified customer relations?

### How to attract consumers?

There should be **benefits for consumers** as well. If the system's cumbersome to
use and most products cost more, no-one will use it except some idealists.

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="law" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Law</h3>
</div>
<div class="panel-body tile-panel-body">

### What are the legal implications of creating an alternative currency?

What are the legal risks? Are there any subtle differences in the handling of
the alternative currency that would influence if it's deemed legal or illegal?

### Transparency vs. secrecy

Are there any **laws protecting company secrets** that would conflict with an open
and transparent economy like in the Viridian project?

Of course, one cannot force companies to reveal information unless one changes
the law, in which case the conflict would be removed. But in an ideal world,
one would manage to encourage companies to reveal information voluntarily until
everyone is doing it and it becomes normal practice.

But still, are there any legal obstacles to the whole process?

### Any other law problems we didn't think about?

Law can be very unintuitive and incomprehensible for laymen. Maybe there is
**an obvious legal obstacle** that did not occur to us yet?

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="lca" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Life-cycle analysis, external costs, sustainability</h3>
</div>
<div class="panel-body tile-panel-body">

### How can the Viridian project benefit from existing LCA work?

Life-cycle analysis (LCA) is the **assessment of environmental impacts** associated
with the entire lifetime of a product, from the extraction of raw materials, via
production and use, to its disposal and recycling.

**Databases and dedicated software products** for LCA already exist and are
regularly updated by researchers. These would provide a solid base for the
determination of many products' colors.

A problem is that existing LCA products like [ecoinvent](http://www.ecoinvent.org/)
are usually **not freely available and open sourced**. Unless there are APIs allowing
to interact with them, it will be rather complicated to use those third party
software products. A very welcome open source software project is [openLCA](http://www.openlca.org/),
which solves some of these problems.
However, another problem is **licensation**, which might even prohibit the use of
the data for such purposes as the Viridian project.

The problem of course is that LCA is a very labor-intensive task and the
researchers doing it must use licensation to ensure that they earn a living.
In the Viridian economy, LCA researchers would turn into color miners that are
paid fair rewards for their services, but only if they open up all their data.
Therefore, they would be in a much different (and much more important and powerful)
position than they are now. The question is: **how to organize the transition** from
here to there?

### Are there better ways to assess external costs?

Apart from LCA, can you think of alternative ways to determine external costs
that are easier to use?

Are there any other tools in the sustainability and external costs scene that
would be really useful for the Viridian project?

### Contact us

If you are an LCA/external costs/sustainability expert and think you can help,
contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="auditing" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Auditing, e.g. organic or fair trade</h3>
</div>
<div class="panel-body tile-panel-body">

### Any learned lessons from auditing practice?

Already existing sustainability certifications of products (fair trade, organic
standards, etc.) usually involve audits, i.e. **systematic and unscheduled visits** to
producers to check if the required standards are met and that there is no abuse
of the certificate label.

Do these kind of audits need to be a part of the color determination? How can
they be organized as **efficient and corruption-safe** as possible? Are there any
**mistakes** that can be avoided? Any thoughts on how to **scale audits up** in a good
way? Or do you think audits are actually a bad thing and there are
**better alternatives**, e.g. using social networks and personal contacts to people?

### Contact us

If you have experience from the auditing business and think you can help,
please contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="sociology" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Psychology and Sociology</h3>
</div>
<div class="panel-body tile-panel-body">

### The rational brain and the emotional brain?

Economic decisions seem to be driven mostly by rational thinking and self-interest
in the lines of **"cheaper is better" and "more is better"**. Negative
effects on others or the environment are usually not considered when people buy
things, or (to a lesser degree) even when they look for an employer.

But people are not just rational, but also social and emotional beings. The
hypothesis is that while **conventional money stimulates mainly the rational part of the brain**,
there could also be **a component of money that stimulates the emotional part**.
With the emotional component putting an emphasis on empathy and
ethics, the perception and use of money could become much more balanced and
less one-dimensional and mechanistic than it is now. The color dimension of the
Viridian money could provide this extra mechanism.

Does this hypothesis have a sound foundation in psychology or is there a stark
**contradiction with psychological research?**

### How to use psychology to make Viridian attractive?

Capitalist advertising makes use of all kinds of psychological tricks to entice
people and encourage them to change their behavior (mostly to buy stuff they
don't really need).

What kind of **psychological tricks** could be used **to convince people to use the ethical Vidirian money?**
Could peer-pressure be used in a positive way to attract new users? Maybe in the
lines of "Three of your friends already use Viridian, when do you come and join us?"

### What about socio-economic aspects?

The socio-economic situation of the people is important because it ultimately
decides about what people perceive as "good and fair" and what they accept and
support. If there is a perceived thread to peoples' lifestyle, there will be
opposition.

Environmental action and climate action is often put on the shoulders of the
weakest. For example, when stricter regulations lead to higher prices, it is
the poor that suffer most from it, while the rich might not even feel a difference.
Therefore, it is **very important to implement sustainable development in a way that improves the socio-economic status**
of the vast majority of people and **especially the poorest**. The rich are the
biggest polluters, so they are also the ones that should pay for the transition
towards a safe world.

How can we design Viridian so that the improvement of the socio-economic status
of most people is guaranteed, while the economy gets more sustainable at the
same time?

### Contact us

Do you have any thoughts, ideas, comments, suggestions, improvements? Contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="politics" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Politics: Democracy and Voting</h3>
</div>
<div class="panel-body tile-panel-body">

### How to organize ourselves?

Politics is all about power and who gets to decide what. Conventional political
systems use **centralized structures for political decision making**. Computer
technology may enable us to use better alternatives.

In the Viridian project, we want **more decentralized and bottom-up** decision
making to minimize corruption and increase societal acceptance. But because
direct democracy can be very inefficient or prone to demagogical manipulation,
a **middle way between direct and representative democracy** might prove very useful.
*Delegative democracy*, also called *liquid democracy*, is a process, in
which political decisions are not delegated completely to just one representative,
but each voter can decide by topic whom to delegate questions to. Or, if the voter
feels competent in the subject matter, she can choose to answer the question herself.

Direct democracy has the general problem that it does not scale well. It works
well in small groups, so that cooperatives, for instance, can be operated very
efficiently. But a whole society, operated like a cooperative, is difficult to
achieve. Liquid democracy might combine the best of both worlds: The direct
involvement of direct democracy and the stability and efficiency of representative
democracy. Liquid democracy involves a lot of communication between people,
making use of social networks. That means, it can profit heavily from the use of
information technology, without which it might be too cumbersome and slow and
therefore inefficient.

### Contact us

If you have any other ideas on how decisions in the Viridian project might become
more democratic and legitimate, or if you have experience with liquid democracy
or similar tools, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>


## How is this any different from what already exists?

There are already many good concepts for a sustainable economy. See how they
compare to the Viridian project.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-comp" aria-expanded="false" aria-controls="showmore-comp"><i class="fa fa-caret-down" aria-hidden="true"></i> Read more</button>

<p></p>

<div class="collapse" id="showmore-comp">

Existing ideas for a more sustainable economy include:

* [Transition Towns](https://transitionnetwork.org/) and [Permaculture](https://en.wikipedia.org/wiki/Permaculture) (see also: [here](https://permacultureprinciples.com/))
* [Degrowth](https://www.degrowth.info)
* [Cradle to cradle](https://en.wikipedia.org/wiki/Cradle_to_Cradle:_Remaking_the_Way_We_Make_Things) (see also: [here](http://c2c-ev.de/))
* [Ecological economics](https://en.wikipedia.org/wiki/Ecological_economics) (see also: [here](https://www.journals.elsevier.com/ecological-economics/))
* [Sustainable Economics](https://www.nachhaltige-oekonomie.de/englisch)
* [*Gemeinwohlökonomie*](https://www.ecogood.org) (German)
* *Öko-Bilanzierung* (German) (["Richtig rechnen!"](https://www.oekom.de/nc/buecher/gesamtprogramm/buch/richtig-rechnen.html))
* [*Bioökonomie*](https://de.wikipedia.org/wiki/Bio%C3%B6konomie) (German)
<!-- * True cost -->

While these are very good approaches, existing concepts for a more sustainable
economy typically lack some or all of the following elements:

<img src="/images/home/coding.svg" style="height: 75px;">

<!--
<div>Icons made by <a href="https://www.flaticon.com/authors/becris" title="Becris">Becris</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div>
-->

#### Tech

While technology alone cannot solve our problems and social
innovation is needed, computer network technology (web, digitization,
cryptocurrencies, IoT, etc.) has a **vast potential to support or even enable societal transformation**.
For example, technology can be critical for allowing
humans to efficiently collaborate and exchange information. It might
also be used to algorithmically enforce sustainability (whether this is feasible
or not remains to be seen).

<img src="/images/home/sunglasses.svg" style="height: 75px; margin-bottom: -25px;">

<!--
<div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div>
-->

#### Coolness


For the alternative technology to be successful, it must be "cool"
enough to be adopted by a large (but minority) group of "early adopters".
If the new technology **"goes viral"**, it will spread very fast and be talked
about a lot. This would be the first stage towards mass adoption.
<!-- In the Daemon novels, this was achieved by addressing the gamer community and -->
<!-- transferring concepts of comupter games to the real world. -->

<!-- <img src="/images/home/stats.svg" style="height: 75px;"> -->

<!--
<div>Icons made by <a href="https://www.flaticon.com/authors/vectors-market" title="Vectors Market">Vectors Market</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div>
-->

<img src="/images/home/coins.svg" style="height: 75px;">

<!--
<div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div>
-->

#### Benefit

The alternative economy will only be widely adopted if it provides
real benefit for the individual. Mere moral superiority is probably not enough.
One example might be farmers: Many farmers need to pay large amounts to corporations
to buy seeds, fertilizers, and pesiticides, while not earning much with their
crops on the market. If they discover that they can **earn a much better living**
by using the alternative economy, both for buying goods and selling their
produce, they will start using it. This would be the second stage towards mass
adoption. When more and more groups of people join the new economy, eventually
a **critical mass** will be reached and there will be a tipping point with a full
transition to the new economy.

The Viridian project does not yet have complete solutions on these three points
either. But we have a vision. And you can help us turn that vision into a
reality! See [Get involved](#get-involved).

</div>

####

<div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a>, <a href="https://www.flaticon.com/authors/retinaicons" title="Retinaicons">Retinaicons</a>, <a href="https://www.flaticon.com/authors/smashicons" title="Smashicons">Smashicons</a>, <a href="https://www.flaticon.com/authors/zlatko-najdenovski" title="Zlatko Najdenovski">Zlatko Najdenovski</a>, <a href="https://www.flaticon.com/authors/becris" title="Becris">Becris</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> are licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a>.</div>
 
```{r load-javascript, echo=FALSE}
htmltools::includeScript("../static/js/tiles.js")
```
