---
#title: Das Viridian-Projekt
#subtitle: Für eine nachhaltige und gerechte Wirtschaft
author: mvoge
date: '2019-03-24'
---

## Für eine nachhaltige und gerechte Wirtschaft

Vieles scheint **falsch zu laufen** in unserer Welt: wachsende soziale Ungleichheit,
ungebremster Klimawandel, Rückgang der Artenvielfalt... um nur einige Probleme zu nennen.
**Bist du auch der Meinung, dass Änderungen notwendig sind, aber nie etwas passiert?**
Dann könnte dieses Projekt für dich interessant sein.

<!--
**Gehörst du auch zu denen, die glauben, dass wir dringend umdenken müssen, aber vergeblich darauf warten, dass sich etwas ändert?**
-->

Die meisten Probleme der Menschheit haben mit Wirtschaft zu tun. Deshalb möchte
das *Viridan-Projekt* die **Art des Wirtschaftens** verändern. Stelle dir eine
Wirtschaft vor, die nicht nur finanzökonomische, sondern auch ökologische und
soziale Belange berücksichtigt. Eine Wirtschaft, in der Menschen mehr Wert sind
als Profite. Eine Wirtschaft, in der **nachhaltig und fair die Regel sind** und
nicht die Ausnahme.

<!--
Die Inspiration für das Viridian-Projekt stammt von zwei Gemeinden, die wenig haben
Kontakt miteinander bisher: Die Vision der * Transition Town * Bewegung von
eine ** nachhaltige, postfossile, belastbare Wirtschaft ** auf einer Seite. Die Vision der
* Kryptowährungsgemeinschaft * einer ** dezentralisierten Peer-to-Peer-Wirtschaft ** auf der anderen Seite
Seite.
-->

Das Hauptziel des Viridian-Projekts besteht darin,
**externe Kosten unter Verwendung einer neuen digitalen Währung zu internalisieren**.
Dadurch werden soziale und ökologische Belange zu einem normalen Bestandteil
jedes Geschäftsmodells.

<!--
Dies schafft Anreize für Unternehmen, die nicht schädlich sind
Gesellschaft. Wenn sozialer Gruppenzwang in die Wirtschaftstätigkeit einbezogen wird (über die
alternativer digitaler Währung) werden die Menschen automatisch verantwortungsbewusster.
Dies ähnelt der Art und Weise, wie * Commons * funktioniert, bei denen eine Gruppe von Personen arbeitet
Kollektiv kümmert sich um eine gemeinsame Ressource, die jeden Einzelnen verantwortlich macht.
-->

<!--
Um dies zu erreichen, soll ein Computernetzwerk implementiert werden, in dem die gewünschten
Anreize sind algorithmisch inhärent und werden kollektiv von der Menge bestimmt.
Diese Algorithmen müssen sehr sorgfältig entworfen, transparent und sein
von allen vereinbart. Sie müssen auch ständig debattiert und angepasst werden
so bald wie nötig.
-->

<a href="/paper" class="btn btn-info" role="button">Lies das Whitepaper (konzeptionell)</a>
<a href="flyer.pdf" class="btn btn-info" role="button">Sieh dir den Flyer an</a>
<a href="technical" class="btn btn-info" role="button">Lies den technischen Artikel</a>

## Ziele

Ziel ist es, ein Computernetz für die gemeinschaftliche Nachhaltigkeitsbewertung
von Produkten und Unternehmen zu schaffen. Dadurch soll es möglich werden, externe
Kosten in die Preise einzubeziehen. Das Netzwerk soll die folgenden Vorzüge aufweisen:

<div markdown class="row">
<div markdown class="col-md-6">

![Dezentralisierung](/images/home/network.svg){style="height: 75px;"}

Das Netzwerk soll dezentralisiert sein für erhöhte **Resilienz**.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-dec" aria-expanded="false" aria-controls="showmore-dec"><i class="fa fa-caret-down" aria-hidden="true"></i> Mehr anzeigen</button>

<!-- <i class="glyphicon glyphicon-triangle-bottom"></i> -->
<!-- &#x25bc; UTF-8 alternative to glyphicon-triangle-bottom -->

<p></p>

<div class="collapse" id="showmore-dec">

Es sollte keine singulären Schwachstellen (engl. "Single Point of Failure") geben,
deren Ausfall schwerwiegende Auswirkungen hätte. Es muss auch verteilt sein, um
**zu starke Machtkonzentration zu vermeiden**. Das Netzwerk soll in den Händen
aller Menschen sein und auch von ihnen betrieben werden, nicht von wenigen
(oder einzelnen) Unternehmen oder Regierungen.

</div>

<!-- <div>Von mir erstelltes Symbol!</div> -->

</div>
<div class="col-md-6">

<div class="figure">
![Demokratisierung](/images/home/like.svg){style="height: 75px;"}
![Demokratisierung](/images/home/dislike.svg){style="height: 75px;"}
<p class="caption">Demokratisierung</p>
</div>

Dank Dezentralisierung und Transparenz werden Wirtschaft und Gesellschaft
**demokratischer und gleichberechtigter**.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-dem" aria-expanded="false" aria-controls="showmore-dem"><i class="fa fa-caret-down" aria-hidden="true"></i> Mehr anzeigen</button>

<p></p>

<div class="collapse" id="showmore-dem">

Wenn das Netzwerk auf vielen dezentralen Knoten ausgeführt wird, kann die Konzentration
von Ressourcen und Macht verhindert werden. Dezentralisierte Systeme wie z.B.
Blockchains ermöglichen eine **Machtverschiebung** weg von zentralisierten Systemen
(z.B. Banken, Konzernen) hin zu den Menschen, die diese Infrastruktur verwenden.
Außderdem sollen die Regeln des Netzwerks, insbesondere die Bewertung von Produkten und
Unternehmen, **von den Nutzern festgelegt werden**: in einem offenen, kollaborativen
Prozess, ähnlich der Entwicklung von Open-Source-Software. Die Verwendung von
Computernetzwerken ermöglicht die Verwendung von innovativen demokratischen
Konsenswerkzeugen wie "Liquid Democracy".

</div>

<!-- <div>Icons made by <a href="https://www.flaticon.com/authors/smashicons" title="Smashicons">Smashicons</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
</div>

####

<div class="row">
<div class="col-md-6">

![Nachhaltigkeit](/images/home/sprout.svg){style="height: 75px;"}

Transparente Preise, inklusive externer Kosten, **setzen Anreize** für ein
nachhaltiges und sozial verantwortliches Wirtschaften.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-sust" aria-expanded="false" aria-controls="showmore-sust"><i class="fa fa-caret-down" aria-hidden="true"></i> Mehr anzeigen</button>

<p></p>

<div class="collapse" id="showmore-sust">

Nachhaltige und sozial verantwortliche Wirtschaft und Handel werden angeregt
**durch die Internalisierung der externen Kosten** wirtschaftlicher Tätigkeit
in den Preisen. Verursacht ein Unternehmen hohe externe Kosten, um dessen
Vermeidung oder Beseitigung es sich nicht kümmert, so bekommt es dadurch oft einen
**unfairen Wettbewerbsvorteil**. Die Kosten fallen aber trotzdem an, für uns alle
(oft den Steuerzahler). Sie werden sozialisiert. Wenn die Preise von vornherein
externe Kosten beinhalten, bekommen Unternehmen einen Wettbewerbsvorteil, wenn sie
sie reduzieren, anstatt Kosten immer weiter zu sozialisieren.
Gesellschaftlich sinnvolle Lösungen werden automatisch bevorzugt.

Darüber hinaus ist es für die Nachhaltigkeit wichtig, dass die Transaktionen der
Blockchain nicht in einem ressourcenintensiven Proof-of-Work-Prozess verifiziert
werden, sondern dass eine Alternative mit einem möglichst kleinen ökologischen
Fußabdruck verwendet wird.

</div>

<!-- <div>Icons made by <a href="https://www.flaticon.com/authors/retinaicons" title="Retinaicons">Retinaicons</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
<div class="col-md-6">

![Fairness](/images/home/handshake.svg){style="height: 75px;"}

Der Viridian-Markt wird fairer sein und **einem idealisierten freien Markt näherkommen**.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-fair" aria-expanded="false" aria-controls="showmore-fair"><i class="fa fa-caret-down" aria-hidden="true"></i> Mehr anzeigen</button>

<p></p>

<div class="collapse" id="showmore-fair">

Wenn externe Kosten in die Preisberechnung einbezogen werden,
ist der Markt fairer und nähert sich einem idealisierten freien Markt an.
Auf einem solchen Markt sind **alle Marktteilnehmer perfekt informiert**. Die Preise,
die sich ergeben, spiegeln die **Kosten** wider, die
**der Gesellschaft durch die Produktion entstanden sind**. Anders gesagt: Jede(r)
von der Produktion Beeinträchtigte wird angemessen entschädigt. Denn wird jemand
nicht angemessen entschädigt, kann er/sie auf einem freien Markt einfach woanders
hingehen. Im Falle von externen Kosten funktioniert das so nicht. Der Markt ist
imperfekt und Viridian schließt diese Lücke ein Stück weit.

Weil unter Umständen die Lieferkette bekannt wird, kann dadurch auch minderwertige
Qualität aufgedeckt werden. Kunden müssen nicht auf den Ruf eines Herstellers vertrauen,
sondern können sich für das beste Angebot entscheiden. Dies erhöht die Chancen
von neuen Konkurrenten, sich mit ihren besseren Produkten durchzusetzen. Insbesondere
wenn dadurch externe Kosten eingespart werden.
Das **beschleunigt die Einführung alternativer Technologien**. Innovation wird
gefördert und Marktmonopole werden verringert.

</div>

<!-- <div>Icons made by <a href="https://www.flaticon.com/authors/zlatko-najdenovski" title="Zlatko Najdenovski">Zlatko Najdenovski</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
</div>

####

<div class="row">
<div class="col-md-6">

![Transparenz](/images/home/search.svg){style="height: 75px;"}

**Probleme in der Lieferkette sollen aufgedeckt werden.** Neben der Bewertung
eines Produktes oder Unternehmens soll auch sichtbar sein, *warum* es diese
Bewertung erhält.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-trans" aria-expanded="false" aria-controls="showmore-trans"><i class="fa fa-caret-down" aria-hidden="true"></i> Mehr anzeigen</button>

<p></p>

<div class="collapse" id="showmore-trans">

Der Bewertungvorgang und -algorithmus sollen möglichst transparent sein. Änderungen
an einer Bewertung oder am Algorithmus müssen von legitimierenden Informationen
bzw. belegbaren Quellen begleitet werden.
**Öffentlichkeitsrelevante Daten müssen für alle zugänglich sein.**
Aufgrund ihrer Bedeutung für die Gesellschaft müssen Informationen über die
externen Kosten einer wirtschaftlichen Tätigkeit öffentlich vorliegen
und können kein Betriebsgeheimnis sein. Weil Menschen wichtiger sind als Profite,
kann es keinen solchen "Anlegerschutz" geben.

</div>

<!-- <div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
<div class="col-md-6">

![Datenschutz](/images/home/lock.svg){style="height: 75px;"}

Das Netzwerk muss **Ende-zu-Ende-verschlüsselt** sein, um Sicherheit und
Privatsphäre zu gewährleisten.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-priv" aria-expanded="false" aria-controls="showmore-priv"><i class="fa fa-caret-down" aria-hidden="true"></i> Mehr anzeigen</button>

<p></p>

<div class="collapse" id="showmore-priv">

Die Menschen sollten die volle Kontrolle über ihre Daten haben.
**Private Daten sollten geheim gehalten werden**, anstatt eine Ware zu sein,
mit der Gewinne erzielt werden.

</div>

<!-- <div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div> -->

</div>
</div>

<!-- * **Transaktionen:** Über das Netzwerk sollen Finanztransaktionen möglich sein, -->
<!--   d,h. Wertmarken (eine Währung) sollen über das Netzwerk ausgetauscht werden. -->


## Mach mit

Im Moment sind wir nur ein kleines Team (siehe [Über das Projekt](about) und [Kontakt](contact)).
Über mehr Unterstützung würden wir uns freuen.

Die anfängliche Konzeptionsphase ist langsam zu Ende, die Vision wird immer konkreter
und wir beginnen mit der Implementierung (siehe [Timeline](technical/#timeline)).
Trotzdem gibt es immer noch viele offene Fragen und ungelöste Probleme.
Deshalb sind wir **immer auf der Suche nach Gleichgesinnten, um zu diskutieren und Ideen auszutauschen.**
Besonders zu den Themen auf den Kacheln unter [Offene Fragen](#offene-fragen)
gibt es noch Gesprächsbedarf.

Konkretere Überlegungen zur Implementierung gibt es im Artikel
[Technisches](technical). Ideen oder Anregungen dazu [nehmen wir gerne entgegen](contact).


## Offene Fragen

Hier sind die verschiedenen Themengebiete und ihre offenen Fragen gelistet.
Wenn du uns bei etwas weiterhelfen kannst, dann melde dich bitte: [Kontakt](contact).

```{r, echo=FALSE}
htmltools::includeHTML("snippets/tiles_de.html")
```

<div style="margin-bottom: 30px;"></div>

<div id="dlt" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Distributed-Ledger-Technologie</h3>
</div>
<div class="panel-body tile-panel-body">

### Which solution to choose?

Bitcoin, Litecoin, Ethereum, IOTA, blockchain, DAG, smart contracts, DAOs,
proof-of-work, proof-of-stake, proof-of-cooperation, proof-of-interest, ... ---
the world of distributed ledgers and cryptocurrencies is full of innovative ideas,
with new ones springing up monthly.

**If you are enthusiastic about all things crypto, then you might be able to help us decide, which solution is best suitable for the Viridian project.**

We want transaction validation to have an as small resource footprint as
possible. The incentive system must be designed such that there is no motivation
for running a lot of computing power, however there must be motivation to
contribute to the network in the long term. Of course, the system shall be as
decentralized and trustless as possible to keep it secure and scalable.

**Please note:** In the meantime we settled to use Hyperledger Fabric, a
permissioned blockchain, as a manipulation-resistant decentralized database
backend. It does not provide as much decentralization and trustlessness as
proof-of-work blockchains like Ethereum, but we believe it to be a good compromise
between resilience and efficiency. See [the technical paper](technical) for more
details.

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="software-development" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Software development</h3>
</div>
<div class="panel-body tile-panel-body">

### Good design is the key

The Viridian project's concept relies heavily on the use of computer technology
and software. All this software still has to be built, both on the backend side
and on the frontend/UI/UX side.

**A clear vision and good design choices are crucial for an infrastructure that is secure and maintainable and that scales well.**

One should build upon existing work wherever possible. Especially when it comes
to cryptography, there is a good rule of thumb that one should not "come up
with a home-brew solution". Therefore, a first step would be, together with
creating the vision, to carefully evaluate which components could be based on
what, and how these components must be able to interact with each other, keeping
in mind maintainability, security and scalability.

Once these general points are sorted out and the roadmap is clear, one can set
up the requirements for the various components and can start implementing them
(perhaps just making adjustments or additions to already existing code).

A slick UI/UX will also be a crucial part of attracting users.

**Please note:** In the meantime we settled to use Hyperledger Fabric, an
open source blockchain framework maintained by the Linux Foundation. The JSON-based
distibuted database CouchDB will be used as the backend of Fabric to store the data.
Therefore, on the backend and API side, we are mostly set and will use the infrastructure
and crpytography provided by Hyperledger Fabric. On the frontend side, we currently
plan to develop a Progressive Web App (PWA), but might also look into Flutter or Cordova.
See [the technical paper](technical) for more details.

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="monetary" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Currencies and monetary policy</h3>
</div>
<div class="panel-body tile-panel-body">

### What are the rules of a good currency?

In traditional fiat currencies, a central bank is in charge of controlling the
currency, e.g. by regulating the amount of money in circulation, or in fact by
setting the key interest rate.

In a decentralized cryptocurrency, there is no central bank, which is a good thing
because it prevents corruption and dependency on centralized entities.
Instead, the network's mathematical algorithms take over the role of a central bank.

An important question is: **How to design these algorithmic rules?** For instance,
what is the **optimal supply of fresh money**? Shall there be a fixed amount of money?
Or shall the amount of money increase (causing inflation) or decrease? Shall
there be a demurrage imposed on the currency, constantly devaluating it over
time?

### What role will the financial system play?

In the Viridian economy, will banks still be important players? Private banks
currently create almost all the money that enters the economy. With a decentralized
cryptocurrency as dominant payment medium, this would not be the case anymore.

So, how shall banks be treated in the Viridian economy? Will banks' services also
have associated colors? **Will banks be superseded** by innovative new concepts, e.g.
peer-to-peer credit systems with risk scattering by spreading investment money
on many agents?

What to do with interest rates on investments? **Shall there still be interest rates?**
Or are interest rates not acceptable anymore since they are drivers of limitless
growth in a world with limited resources?

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="economics" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Economics</h3>
</div>
<div class="panel-body tile-panel-body">

### Would the Viridian economy be closer to an "ideal" free market?

The concept of *free market economy* goes back to the influential economists
Adam Smith and David Ricardo. Smith introduced the idea that a free market
results in fair conditions for all participants. Ricardo formalized and abstracted
Smith's ideas and had to simplify it a lot. This resulted in a well working
theory, but many people also started to neglect the limitations of the model
and the fact that it does not necessarily represent reality.

Adam Smith was very aware of the limitations of his concept and warned about them.
For instance, a market with few corporations is a bad idea because the corporations
can choose to require higher prices than they need to. It becomes even worse when
big business and government make each other gifts, abusing their power, leading
to high profits for few capitalist leaders and corrupt politics. This is why
Smith was a fierce opponent of corporations and preferred an economy with many
small competitors (like most of economy still was at Smith's time, which changed
over the course of the 19th century).

The Viridian economy, with its decentralized power and easily adjustable rules,
**might help to prevent such situations and create a market that is closer to Smith's ideal**.

Another flaw of our present-day capitalist economy is the exclusion of external
costs. Many economists argue that this leads to a distortion of the free market.
**In Smith's ideal free market, all prices reflect the actual cost that a product's production imposes on society.**
For example, if a worker needs to spend more time on one product than on another,
then it must cost more to reflect this. If the production leads to external costs
that society (ultimately everyone, including the buyer and the seller) must pay for,
then the product should also cost more, to pay for these costs and also to regulate
the demand for the product.

Again, the Viridian project might help to remove this flaw from the economy.

What do you think about it? If you have any opposing views or ideas how the
Viridian project can improve our market economy even more, feel free to contact
us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

### What are the likely conerns and needs of businesses?

The Viridian project would necessarily interfere with the way people are doing
business. For example, color must be considered when determining retail prices.

From the business perspective, what would the main **concerns and objections** be
that might disturb the adoption of Viridian? How could these concerns be addressed?
What are the **possibilities to make a transition easier for businesses**? Can we
provide services that would simplify a business's life?

### How to attract businesses?

How can we create business models that would
**make the adoption of Viridian lucrative for businesses**,
e.g. through improved image or intensified customer relations?

### How to attract consumers?

There should be **benefits for consumers** as well. If the system's cumbersome to
use and most products cost more, no-one will use it except some idealists.

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="law" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Law</h3>
</div>
<div class="panel-body tile-panel-body">

### What are the legal implications of creating an alternative currency?

What are the legal risks? Are there any subtle differences in the handling of
the alternative currency that would influence if it's deemed legal or illegal?

### Transparency vs. secrecy

Are there any **laws protecting company secrets** that would conflict with an open
and transparent economy like in the Viridian project?

Of course, one cannot force companies to reveal information unless one changes
the law, in which case the conflict would be removed. But in an ideal world,
one would manage to encourage companies to reveal information voluntarily until
everyone is doing it and it becomes normal practice.

But still, are there any legal obstacles to the whole process?

### Any other law problems we didn't think about?

Law can be very unintuitive and incomprehensible for laymen. Maybe there is
**an obvious legal obstacle** that did not occur to us yet?

### Contact us

If you think you can help, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="lca" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Life-cycle analysis, external costs, sustainability</h3>
</div>
<div class="panel-body tile-panel-body">

### How can the Viridian project benefit from existing LCA work?

Life-cycle analysis (LCA) is the **assessment of environmental impacts** associated
with the entire lifetime of a product, from the extraction of raw materials, via
production and use, to its disposal and recycling.

**Databases and dedicated software products** for LCA already exist and are
regularly updated by researchers. These would provide a solid base for the
determination of many products' colors.

A problem is that existing LCA products like [ecoinvent](http://www.ecoinvent.org/)
are usually **not freely available and open sourced**. Unless there are APIs allowing
to interact with them, it will be rather complicated to use those third party
software products. A very welcome open source software project is [openLCA](http://www.openlca.org/),
which solves some of these problems.
However, another problem is **licensation**, which might even prohibit the use of
the data for such purposes as the Viridian project.

The problem of course is that LCA is a very labor-intensive task and the
researchers doing it must use licensation to ensure that they earn a living.
In the Viridian economy, LCA researchers would turn into color miners that are
paid fair rewards for their services, but only if they open up all their data.
Therefore, they would be in a much different (and much more important and powerful)
position than they are now. The question is: **how to organize the transition** from
here to there?

### Are there better ways to assess external costs?

Apart from LCA, can you think of alternative ways to determine external costs
that are easier to use?

Are there any other tools in the sustainability and external costs scene that
would be really useful for the Viridian project?

### Contact us

If you are an LCA/external costs/sustainability expert and think you can help,
contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="auditing" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Auditing, e.g. organic or fair trade</h3>
</div>
<div class="panel-body tile-panel-body">

### Any learned lessons from auditing practice?

Already existing sustainability certifications of products (fair trade, organic
standards, etc.) usually involve audits, i.e. **systematic and unscheduled visits** to
producers to check if the required standards are met and that there is no abuse
of the certificate label.

Do these kind of audits need to be a part of the color determination? How can
they be organized as **efficient and corruption-safe** as possible? Are there any
**mistakes** that can be avoided? Any thoughts on how to **scale audits up** in a good
way? Or do you think audits are actually a bad thing and there are
**better alternatives**, e.g. using social networks and personal contacts to people?

### Contact us

If you have experience from the auditing business and think you can help,
please contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="sociology" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Psychology and Sociology</h3>
</div>
<div class="panel-body tile-panel-body">

### The rational brain and the emotional brain?

Economic decisions seem to be driven mostly by rational thinking and self-interest
in the lines of **"cheaper is better" and "more is better"**. Negative
effects on others or the environment are usually not considered when people buy
things, or (to a lesser degree) even when they look for an employer.

But people are not just rational, but also social and emotional beings. The
hypothesis is that while **conventional money stimulates mainly the rational part of the brain**,
there could also be **a component of money that stimulates the emotional part**.
With the emotional component putting an emphasis on empathy and
ethics, the perception and use of money could become much more balanced and
less one-dimensional and mechanistic than it is now. The color dimension of the
Viridian money could provide this extra mechanism.

Does this hypothesis have a sound foundation in psychology or is there a stark
**contradiction with psychological research?**

### How to use psychology to make Viridian attractive?

Capitalist advertising makes use of all kinds of psychological tricks to entice
people and encourage them to change their behavior (mostly to buy stuff they
don't really need).

What kind of **psychological tricks** could be used **to convince people to use the ethical Vidirian money?**
Could peer-pressure be used in a positive way to attract new users? Maybe in the
lines of "Three of your friends already use Viridian, when do you come and join us?"

### What about socio-economic aspects?

The socio-economic situation of the people is important because it ultimately
decides about what people perceive as "good and fair" and what they accept and
support. If there is a perceived thread to peoples' lifestyle, there will be
opposition.

Environmental action and climate action is often put on the shoulders of the
weakest. For example, when stricter regulations lead to higher prices, it is
the poor that suffer most from it, while the rich might not even feel a difference.
Therefore, it is **very important to implement sustainable development in a way that improves the socio-economic status**
of the vast majority of people and **especially the poorest**. The rich are the
biggest polluters, so they are also the ones that should pay for the transition
towards a safe world.

How can we design Viridian so that the improvement of the socio-economic status
of most people is guaranteed, while the economy gets more sustainable at the
same time?

### Contact us

Do you have any thoughts, ideas, comments, suggestions, improvements? Contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>

<div id="politics" class="anchor">
<div class="panel panel-default">
<div class="panel-heading tile-panel-heading">
<h3 class="panel-title">Politics: Democracy and Voting</h3>
</div>
<div class="panel-body tile-panel-body">

### How to organize ourselves?

Politics is all about power and who gets to decide what. Conventional political
systems use **centralized structures for political decision making**. Computer
technology may enable us to use better alternatives.

In the Viridian project, we want **more decentralized and bottom-up** decision
making to minimize corruption and increase societal acceptance. But because
direct democracy can be very inefficient or prone to demagogical manipulation,
a **middle way between direct and representative democracy** might prove very useful.
*Delegative democracy*, also called *liquid democracy*, is a process, in
which political decisions are not delegated completely to just one representative,
but each voter can decide by topic whom to delegate questions to. Or, if the voter
feels competent in the subject matter, she can choose to answer the question herself.

Direct democracy has the general problem that it does not scale well. It works
well in small groups, so that cooperatives, for instance, can be operated very
efficiently. But a whole society, operated like a cooperative, is difficult to
achieve. Liquid democracy might combine the best of both worlds: The direct
involvement of direct democracy and the stability and efficiency of representative
democracy. Liquid democracy involves a lot of communication between people,
making use of social networks. That means, it can profit heavily from the use of
information technology, without which it might be too cumbersome and slow and
therefore inefficient.

### Contact us

If you have any other ideas on how decisions in the Viridian project might become
more democratic and legitimate, or if you have experience with liquid democracy
or similar tools, contact us at
```{r, echo=FALSE}
htmltools::includeHTML("snippets/email.html")
htmltools::includeHTML("snippets/alternative_contact.html")
```

</div>
</div>
</div>


## Gibt es das nicht alles schon?

Es gibt bereits viele gute Ideen und Konzepte für ein nachhaltige Wirtschaft.
Trotzdem glauben wir, dass Viridian gebraucht wird und eine Lücke schließt.

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#showmore-comp" aria-expanded="false" aria-controls="showmore-comp"><i class="fa fa-caret-down" aria-hidden="true"></i> Weiterlesen</button>

<p></p>

<div class="collapse" id="showmore-comp">

Es gibt schon so viele Initiativen zur Nachhaltigkeit und obwohl uns viele bekannt
sind, ist das wahrscheinlich nur ein kleiner Teil. Viele dieser Initiativen
haben uns mehr oder weniger stark beeinflusst.

Zum einen gibt es allgemeine Konzepte zu einer Wirtschaftsweise, die langfristig
sinnvoll ist und mehr in Kreisläufen als linearem quantitativen Wachstum denkt:

- Postwachstum oder englisch [Degrowth](https://www.degrowth.info)
- [Die Ideen von Bernard Lietaer](http://www.lietaer.com/2012/05/money-and-sustainability/) zur Reform des
  Finanzsystems
- [Ökologische Ökonomie](https://en.wikipedia.org/wiki/Ecological_economics) (siehe auch: [hier](https://www.journals.elsevier.com/ecological-economics/))
- [Nachhaltige Ökonomie](https://www.nachhaltige-oekonomie.de) und [Große Transformation](https://www.wbgu.de/hauptgutachten/hg-2011-transformation/)
- [Cradle to cradle](https://en.wikipedia.org/wiki/Cradle_to_Cradle:_Remaking_the_Way_We_Make_Things) (siehe auch: [hier](http://c2c-ev.de/))
- [Transition Towns](https://transitionnetwork.org/) und [Permakultur](https://en.wikipedia.org/wiki/Permaculture) (siehe auch: [hier](https://permacultureprinciples.com/))

Diese Ideen sind aber meist eher philosophischer Natur und geben wenig
konkrete Handlungsvorschläge (wie bei Postwachstum oder Ökologischer Ökonomie)
oder sie sind sehr begrenzt auf bestimmte Sektoren (hauptsächlich Landwirtschaft)
und kleinräumige Gebiete (wie bei Permakultur und Transition Towns).

Konkreter und allgemeiner sind folgende Ansätze, mit denen Nachhaltigkeit fest
in der Unternehmenskultur verankert werden soll, z.B.:

- Die [**Gemeinwohlökonomie**](https://www.ecogood.org) möchte, dass Unternehmen neben
  der herkömmlichen (finanziellen) Bilanz eine "Gemeinwohl-Bilanz"
  erstellen, in die ethische, soziale und ökologische Kriterien eingehen.
- Einen Schritt weiter geht die vorgeschlagene **Reform der Finanzbuchhaltung**
  (["Richtig rechnen!"](https://www.oekom.de/nc/buecher/gesamtprogramm/buch/richtig-rechnen.html))
  von Christian Hiß: Anstatt öko-soziale Belange nur separat von den Finanzen zu
  betrachten, soll beides in einer einzigen Bilanz zusammengeführt werden. Denn
  auf lange Sicht schlagen sich auch ökologisch-soziale Gewinne und Verluste
  finanziell nieder, bzw. sind mit erhöhten oder verminderten unternehmerischen
  Risiken verbunden und müssen daher vom Unternehmer mit berücksichtigt werden.

Diese Ansätze sind sehr zu begrüßen, haben jedoch den Nachteil, dass sie auf
freiwillige Teilnahme der Unternehmen angewiesen sind, solange es keine Verpflichtung
von staatlicher Seite gibt. Da eine politische Mehrheit für eine Verpflichtung
auf Grund der Machtverhältnisse im Status Quo auf absehbare Zeit unwahrscheinlich
ist, geht Viridian einen anderen Weg und setzt auf die Macht der Konsument*innen,
um die Macht der Mächtigen zu brechen.

Zuguterletzt gibt es bereits Initiativen, die ähnlich wie Viridian auf die
Information der Vebraucher*innen zielen und ebenfalls digitale Techniken wie
Datenbanken und Web-Plattformen nutzen:

- Unternehmens- und CSR-Datenbanken:
    - [WikiRate](https://wikirate.org)
    - [CSRHub](https://www.csrhub.com/)
    - [Rank a Brand](https://www.rankabrand.org/)
    - [Corporate Human Rights Benchmark](https://www.corporatebenchmark.org/)
- Siegeldatenbanken:
    - [Siegelklarheit](https://www.siegelklarheit.de/home)
    - [Kompass Nachhaltigkeit](https://www.kompass-nachhaltigkeit.de)
    - [NABU Siegel-Check](https://www.nabu.de/umwelt-und-ressourcen/oekologisch-leben/essen-und-trinken/bio-fair-regional/labels/16627.html)
    - [CIR Siegel von A-Z/Wegweiser durch das Label-Labyrinth](https://www.ci-romero.de/kritischer-konsum/siegel-von-a-z/)
    - [Label-online](https://label-online.de/)
- Produktdatenbanken:
    - [Questionmark](https://www.thequestionmark.org/)
    - [WeGreen](https://web.archive.org/web/20181213124710/http://wegreen.de/) (offline)
    - [Buycott](https://www.buycott.com/)
    - [OpenFoodFacts](https://world.openfoodfacts.org/)
    - [Codecheck](https://www.codecheck.info/)
    - [ToxFox](www.bund.net/toxfox)

In [einem Vergleich](technical/#vergleich) haben wir die Vor- und Nachteile einiger
dieser Initiativen aufgeschrieben und erläutern, was ihnen im Vergleich zu Viridian
fehlt. Außerdem führen wir in unserem technischen Artikel aus, warum das
Viridian-Netzwerk im Gegensatz zu den bestehenden Initiativen sowohl
[gemeinschaftlich bearbeitet](technical/#warum-gemeinschaftlich) als auch
[dezentral organisiert](technical/#warum-dezentral) sein soll.

</div>

####

<div>Icons made by <a href="http://www.freepik.com" title="Freepik">Freepik</a>, <a href="https://www.flaticon.com/authors/retinaicons" title="Retinaicons">Retinaicons</a>, <a href="https://www.flaticon.com/authors/smashicons" title="Smashicons">Smashicons</a>, <a href="https://www.flaticon.com/authors/zlatko-najdenovski" title="Zlatko Najdenovski">Zlatko Najdenovski</a>, <a href="https://www.flaticon.com/authors/becris" title="Becris">Becris</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> are licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a>.</div>
 
```{r load-javascript, echo=FALSE}
htmltools::includeScript("../static/js/tiles.js")
```
