---
title: "The Viridian Project: A digital currency to internalize external costs for a more sustainable economy"
author: "Markus Voge"
date: "`r Sys.Date()` --- v1"
bibliography: references.bib
csl: elsevier-with-titles-alphabetical.csl
# http://editor.citationstyles.org/searchByName/
output: 
  html_document: 
    fig_caption: yes
    number_sections: yes
    self_contained: no
---

```{r functions, include=FALSE}
# A function for captioning and referencing images
# https://stackoverflow.com/questions/13848137/figure-captions-references-using-knitr-and-markdown-to-html
# fig <- local({
#     i <- 0
#     ref <- list()
#     list(
#         cap=function(refName, text) {
#             i <<- i + 1
#             ref[[refName]] <<- i
#             paste("Figure ", i, ": ", text, sep="")
#         },
#         ref=function(refName) {
#             ref[[refName]]
#         })
# })
# Easier solution: captioner package
# http://derekogle.com/fishR/2015-09-17-Figure-Table-Captions-in-Markdown
library(captioner)
figs <- captioner(prefix="Figure")
tbls <- captioner(prefix="Table")
``` 

# Introduction

Among the biggest problems of humanity in the 21st century areenvironmental crises
such as global warming, but also social crises such as inequal distribution of wealth,
with some still suffering from hunger. These problems in turn are catalysts for conflicts
and migration, putting entire societies under stress. Many experts say that we are
pushing the Earth's systems beyond their planetary boundaries [@Rockstrm2009; @Barnosky2012],
and that the current lifestyle of the Global North would require more than one planet to
survive in the long term.[^overshoot]

[^overshoot]: The Global Footprint Network annually calculates the date of the
*Earth Overshoot Day*, on which humanity has depleted the amount of natural resources
that is regenerated each year, see e.g. the 2016 annual report [@GFN2016],
https://www.footprintnetwork.org/our-work/earth-overshoot-day/, or https://www.overshootday.org/.
In 2017, the Earth Overshoot Day fell on 2 August.

Most of these problems have to do with economic activity and the underlying monetary
systems [@Lietaer2012]. One well-known problem is the privatization of profits, while
costs, e.g. of  environmental damages, are being socialized [@wiki:Externality; @wiki:Lemon_socialism].
As a result, often small elites are becoming wealthier, while a majority's well-being is
stagnating or even declining. Another result is the acceptance (also by the suffering majority)
of environmental damage as an inevitable consequence of economic activity and progress,
even when it is risking to destroy the very foundation of life and when healthier
alternatives are both available and financeable by society [see @Klein2014 pages 12ff., 113ff., 117ff.].
One problem of established monetary systems is the inevitability of unlimited economic
growth, which is, as many have pointed out [@Meadows1974; @Georgescu-Roegen1971; @Schneider2010; @Hopkins2014],
opposed to the fact of living on a planet with limited resources. The economic growth
is largely driven by the financial sector and the habit of paying interest rate and
return of investment: companies acquiring money are forced to grow to reel in more
than they invested and pay back their donors. For instance, this drives fossil fuel
companies to ever riskier extraction practices, such as fracking, deep sea drilling,
or drilling in the Arctic, with sometimes catastrophic consequences [see @Klein2014 pages 146--48, 332f., 144].

# Changing the monetary system

As turned out by Belgian economist Bernard Lietaer [@Lietaer2012], all of these
problems might be addressed by monetary innovations. One interesting idea is the
concept of complementary currencies: Currencies that have limited validity, e.g.
only in a certain geographic area or only within certain branches of economy.
These currencies are not to replace national or international currencies, but to
complement them and thereby strengthen closed circles and reduce extreme aggregation
of wealth. The idea of complementary currencies is already actively followed by
people from several groups, e.g. the Transition Towns movement [@Hopkins2014] that
aims to build a post-fossil, sustainable, and resilient economy and by the degrowth
movement [@Douthwaite2012; @Dittmer2013; @Hornborg2017] that aims to put an end to
unlimited economic growth. Complementary currencies are already in use at some places
and prove to strengthen regional and more sustainable value generation. In addition,
there are more ideas to redesign our monetary systems, e.g.
*The Chicago Plan Revisited* [@wiki:The_Chicago_Plan_Revisited; @IMF2012],
a report issued by the International Monetary Fund (IMF) in 2012, suggesting that
private banks should only be allowed to lend money that they have in deposits
(100% reserve banking), reducing both public and private debt and resulting in a
more balanced and stable economy. Another idea is to establish a complementary
currency on the national level in each country, distributed to all citizens as a
basic income, thereby "insulat[ing] local sustainability and resilience from the
deleterious effects of globalization and financial speculation" [@Hornborg2017].

Another monetary innovation are cryptocurrencies (the most popular being Bitcoin)
based on a distributed ledger or blockchain, which do not require any kind of central
institution like a bank [@Nakamoto]. Besides the benefit of reduced cost, a much
more fascinating aspect of cryptocurrencies is that of the democratization of money.
Instead of restricting the privileges of the creation of new money and the verification
of money transactions to few centralized (mostly private) banks, giving them a great
deal of power, this power is distributed by giving it to a large group of relatively
equal players, which everyone can join. A downside is that cryptocurrencies can only
exist digitally and thus require the use of electronic devices. This creates a vulnerable
dependency on electricity and might increase resource consumption,[^crypto_consumption]
which is not good from a sustainability perspective. On the other hand, smartphone
devices capable of digital payments are already now almost ubiquitous, even in developing
parts of the world.

[^crypto_consumption]: In addition to increased use of electronic devices for performing
payments using cyrpto-wallet apps, an enormous energy and resource consumption can be
generated when transactions are verified in a cryptocurrency's network using
so-called *proof-of-work* computations. The Bitcoin protocol is based on proof-of-work
and the Bitcoin network's global energy consumption is assumed to exceed that
of many European countries, including Ireland [@BitcoinEnergy]. There are some
ideas on how to replace proof-of-work with less resource-intensive concepts, e.g.
the *proof-of-stake* used by the Ethereum blockchain.

Cryptocurrencies are not controlled by central institutions, banks and/or governments,
i.e. individual people, but by mathematic algorithms carried out by computer software,
distributed on all participants' devices. This offers several opportunities:
In principle, these algorithms can be formulated in any imaginable way, enabling
the easy implementation of a very different monetary system. Also, if designed properly,
it might be much harder to break the rules of the currency system, since computers,
unlike people, are not corruptible. Especially if manipulation on a majority of
devices would be necessary, then this is virtually impossible, which is also what
makes Bitcoin a very secure currency.


## A new ethical cryptocurrency

The idea is to combine the aim of a more sustainable economy with the benefits of
cryptocurrencies. If cryptocurrencies can be designed as one sees fit, then it might
be possible to solve the problems of conventional monetary systems.

As a starting point, one can try to incorporate external costs into prices of
goods and services. External costs are those that are not part of the economic
calculation when a price is determined, however they still arise (sometimes at a
different place and/or time, harming third parties) and have to be paid by someone,
usually the tax payer or residents of affected areas. Examples of external costs
are health damages caused by pollution or costs of disasters that are the result
of careless extraction of resources, including more frequent extreme weather events
as a result of green house gas emissions. In order to ensure that external costs
are taken care of as well as discourage harmful and encourage sustainable behavior,
it is better to internalize external costs into the actual price.


## Two-dimensional money

The problem is that a (traditional) currency system cannot directly influence
the way a business owner determines prices. The currency as a whole can be valued
up or down by increasing or reducing the amount of fresh money (i.e. setting the
key interest rate), but that does not influence individual trades. However,
cryptocurrencies basically consist of transactions between peers and may offer
a possibility to interfere right there. A possible idea is to add another dimension
to money.

Ever since the invention of money, it has been one-dimensional in its very nature,
the single dimension being value, or rather quantity. Money is a single scale,
against which almost every thing can be measured. Digital currencies might remove
this restriction and could add other dimensions to quantity, for example quality.
Not only the mere amount of (physical) wealth would matter, but also the change
in quality of life that comes along with the aggregation of wealth.

To make it more visual, one can think of the new dimension as color:[^colored_coins]
From green, indicating good quality, to red, indicating bad quality. For means of
computation, color can be converted to a number between -1 (red) and +1 (green).
Since it would be complicated to handle real 2D money (e.g. adding or subtracting
money involves vector calculus, which is simple, but not convenient for many people),
the color should only come into play during a transaction, on the receiver's side.
When the money enters a wallet, it is automatically projected onto a single axis.
This means that red money is valued down, while green money is valued up, which
corresponds to a diagonal movement in the 2D money space. Figures
`r figs("trans-conv", display = "num")` and `r figs("trans-col", display = "num")`
visualize the movement in 2D money space for conventional and colorized
transactions.

[^colored_coins]: Within the cryptocurrency community, already since 2013 there
exists the term "colored coins" (see e.g. http://coloredcoins.org/, https://www.coinprism.com/),
which has nothing to do with the concept described here. "Colored coins" are
bitcoins carrying certain information. They can be used for example to prove
ownership of a certain asset. It is called "colored" because a bitcoin is
"marked" to make it distinguishable from all other bitcoins. Since the ownership
is linked to a cryptocurrency coin, it can be easily transferred to another acount.
This enables the realization of "smart contracts", where digital or real world
assets can be sold securely using a blockchain. Colored coins are similar to
Ethereum, however not using a seperate blockchain, but the bitcoin blockchain.

```{r trans-conv, echo=FALSE, message=FALSE, warning=FALSE, fig.cap=figs("trans-conv", "Conventional money transactions are completely one-dimensional. In this simple example, a seller earns first 1000, then 700 coins.")}
# fig$cap("transactions-conventional", "Conventional money transactions are completely one-dimensional.")
library(fields)
library(grImport)
library(igraph)

# https://stackoverflow.com/questions/28042220/how-to-make-custom-plot-symbols-from-vector-graphics-in-r
PostScriptTrace("coin.eps") # creates .xml in the working directory
coin <- readPicture("coin.eps.xml")

y <- c(0, 1000, 1700)
x <- rep(0, length(y))
plot(x, y, bty="n", xlab = "Color c (Quality)",
     ylab = "Seller's yield y (Quantity)", xlim = c(-1, 1),
     ylim = c(-374, 2238)) # ylim = (max(y) - min(y)) * c(-0.187, 1.119)
xx = grconvertX(x = x, from = 'user', to = 'ndc')
yy = grconvertY(y = y, from = 'user', to = 'ndc')
grid.symbols(coin, x = xx, y = yy, size = 0.05)
# par(new=TRUE)
# plot(c(-1, 1), c(0, 0), axes = FALSE, xlab = "", ylab = "")
colorbar.plot(0, -300, seq(-1, 1, 0.01), strip.length = 1.7, # 0, -0.149 * (max(y) - min(y))
              col = colorRampPalette(c("#990000", "#009900"))(256))
# https://kbroman.wordpress.com/2012/10/11/curved-arrows-in-r/
invisible(igraph:::igraph.Arrows(x[1], y[1], x[2], y[2], sh.lwd = 2,
                                 sh.col = "gray")) # width = 1, size = 1, 
invisible(igraph:::igraph.Arrows(x[2], y[2], x[3], y[3], sh.lwd = 2,
                                 sh.col = "gray"))
text(0.15, 0.5*(y[1] + y[2]), "+1000")
text(0.1, 0.5*(y[2] + y[3]), "+700")
# https://stackoverflow.com/questions/30658177/r-plot-with-axes-in-the-center
# arrows(-1, 0, 1, 0, lwd=1, length=0.15)
# arrows(0, 0, 0, 15, lwd=1, length=0.15)
```

```{r trans-col, echo=FALSE, message=FALSE, warning=FALSE, fig.cap=figs("trans-col", "The same example as in the previous figure, but colorized. With colorized transactions, money transactions are two-dimensional: When a seller obtains money, the transaction's color shifts the value to the right or left. A diagonal projection onto the value axis (y-axis) follows, leading to an increased or decreased value.")}

# y <- c(0, 2000,  2000, 1500, 1000, 1000, 1300, 300,  300, 600)
# x <- c(0,    0, -0.25,    0,    0,  0.3,    0,   0,  0.5,   0)
y <- c(0, 1000,  1000, 1300, 2000, 2000, 1650)
x <- c(0,    0,   0.3,    0,    0, -0.5,    0)
plot(x, y, bty="n", xlab = "Color c (Quality)",
     ylab = "Seller's yield y (Quantity)", xlim = c(-1, 1),
     ylim = c(-374, 2238)) # ylim = (max(y) - min(y)) * c(-0.187, 1.119)
xx = grconvertX(x = x, from = 'user', to = 'ndc')
yy = grconvertY(y = y, from = 'user', to = 'ndc')
grid.symbols(coin, x = xx, y = yy, size = 0.05)
# par(new=TRUE)
# plot(c(-1, 1), c(0, 0), axes = FALSE, xlab = "", ylab = "")
colorbar.plot(0, -300, seq(-1, 1, 0.01), strip.length = 1.7, # 0, -0.149 * (max(y) - min(y))
              col = colorRampPalette(c("#990000", "#009900"))(256))
# https://kbroman.wordpress.com/2012/10/11/curved-arrows-in-r/
# Type head(igraph:::igraph.Arrows) to get all (undocumented) arguments of this internal function
invisible(igraph:::igraph.Arrows(x[1], y[1], x[2], y[2], sh.lwd = 2,
                                 sh.col = "gray")) # width = 1, size = 1, 
invisible(igraph:::igraph.Arrows(x[2], y[2], x[3], y[3], sh.lwd = 2,
                                 sh.col = "#009900"))
invisible(igraph:::igraph.Arrows(x[3], y[3], x[4], y[4], sh.lwd = 2,
                                 sh.col = "#009900"))
invisible(igraph:::igraph.Arrows(x[4], y[4], x[5], y[5], sh.lwd = 2,
                                 curved = -0.3e-3, sh.col = "gray"))
invisible(igraph:::igraph.Arrows(x[5], y[5], x[6], y[6], sh.lwd = 2,
                                 sh.col = "#990000"))
invisible(igraph:::igraph.Arrows(x[6], y[6], x[7], y[7], sh.lwd = 2,
                                 sh.col = "#990000"))
# curve=-0.3e-3
text(0.15, 0.5*(y[1] + y[2]), "+1000")
text(0.2, 1.1*0.5*(y[3] + y[4]), "+300", col = "#009900")
text(0.15, 0.5*(y[4] + y[5]), "+700")
text(-0.25, 0.9*0.5*(y[6] + y[7]), "-350", col = "#990000")
```

*(Icon made by <a href="https://www.flaticon.com/authors/smashicons" title="Smashicons">Smashicons</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a>.)*

It can be argued how exactly this projection should be done, e.g. in a linear or
non-linear way. It is also debatable if there should be upper and lower limits to
the de-/revaluation and what these limits should be. To avoid too extreme value
changes, the effect of the color on the transaction should be considered when the
color is determined. For now, a very simplistic choice is made: A color $c$ of a
transaction changes the value of the reserved money by factor $1 + c$, which can
lie anywhere between 0 (dark red, complete devaluation, i.e. selling is prohibited)
and 2 (dark green, i.e. duplication of value). This means that there is also
"neutral money", which is yellow, $c = 0$, that does not change value, $1 + c = 1$.
So, the price $p$ paid by the buyer is changed into the seller's yield $y$ via:

$$ y = (1 + c) \; p $$

It is also an interesting question whether there should be more up or more down
valuation, which would essentially either create or destroy money, or whether they
should balance each other. This could be controlled via a regularly changing overall
normalization of the re-evaluation, so that the average change in the total amount of
money is constant. Currently in use mainstream currencies constantly create new
money via credits, which creates a low inflation rate. Bitcoin on the other hand,
the most prominent cryptocurrency, has a fixed volume of 21 million bitcoins.
This limited amount of money leads to a deflationary currency. Mainstream economists
argue that it is beneficial to have a low inflation rate, because it prevents people
hoarding money and creates an incentive to spend the money and bring it back into
circulation. This might lead to a more equal distribution of wealth. On the other hand,
it might lead to unnecessary consumption of goods and thus be a driver to (excess)
economic growth. Degrowth economists therefore might favor a deflationary currency.
There are claims that interest rate (and therefore inflation) incentivizes short-term
investments, while an interest-free currency system with demurrage (money losing
some if its value over time, i.e. negative interest rate, or money "disappearing")
incentivizes long-term thinking and promotes a more sustainable economy [@Lietaer2012].

### An example

As a concrete example for a transaction in the new currency, buying an emission-intensive
car with a fossil fuel burning engine might involve red money, while an emission-free
electric vehicle is bought with green money. 

Let's assume that the seller of the red money car had to pay $9000$ coins to buy
the car from the producer. The car seller needs a profit margin of $1000$ coins,
so would sell it for $10000$ coins without considering color. The color of the
transaction is $-0.5$, meaning that the yield $y$ will come out 50% lower than the
price $p$. For $y$ to be $10000$, $p$ must be $p = y / (1 + c)$, so the car must
be sold for $20000$ coins.

On the other hand, the green car seller might have paid $15000$ coins to the
producer and wants to earn $1000$ coins as well. The color is $0.1$, translating
into a price of $16000/1.1 = 14545.45$ coins, making the car cheaper for the
consumer even though it was more costly to produce.

Upon entering the seller's wallet, the red money loses some of its value, while
the green money is valued up. This effectively means that the fossil fuel car will
be sold with a higher price (since the seller gains less), while the electric car
gets cheaper. It is like a higher tax on one, while subsidizing the other. The
buyer does not even need to know about the color of the transaction or the process
of (de-)valuation, only the seller needs to include the color into their price
calculation. This makes it easier to adopt for the consumer. This can help cleaner,
emerging technologies, sometimes needed as quickly as possible from an ecological
point of view, to become marketable much quicker. It can create the right incentives
to make good investments and use good practices, while punishing behavior that is
harmful to society.

It is important to note that, similar to value added tax, the colorization should
probably not be applied multiple times, but only at the end consumer. Intermediate
reevaluations must be undone accordingly.

## Redistribution of value

All transactions regardless of color should include a small transaction fee that
is paid to the color determiners to reward them for their services.

The red money transactions may take a part of the buyer's money and pay it into
a fund to alleviate side-effect damages like environmental disasters or increased
health costs and can also fund dedicated projects supporting unemployed people
during structural change. This would adhere to the principle of "the polluter pays"
[see @Klein2014 pages 110--19] and would be considered fair by most people.

If one introduces demurrage (s.a.), one could also choose not to simply "destroy"
the money, but to redistribute the demurrage evenly across all accounts. This
would act similar to a property tax that reduces inequality.

## Managed by the crowd

While a similar effect to that of colored money can be reached by taxation and
subsidization, the interesting aspect is that computing technology offers a possibility
to do this without a central authority involved. The main problem of transforming
the economy towards sustainable development is to do it in a way that is fair
and socially accepted by everyone. 

Past policies have failed to do so. The less wealthy will not accept a change
that feels like they have to give up their standard of living, while the wealthiest,
the ones most responsible, e.g. those profiting from extraction of resources, are not
paying their toll. The fear of losing jobs due to a necessary system change is
quickly sparked in the public and surely must be addressed, e.g. by dedicated projects.
But because governance fails to introduce such projects, which should be paid for
by polluters like the fossil fuel companies, system change never comes.

This is why such important decisions like the color of money cannot be placed
into the hand of a few privileged, or a single person. It must be done
collectively by the crowd, in a way that is acceptable by everyone.

This is an attemt to bypass government and any centralized institutions or elites:
Like the Bitcoin miners perform the service of validating transactions and keeping
the currency's network secure, dedicated decentralized groups could determine the
transaction colors. Much like developers of open source software, there could be
groups of people dedicated to auditing certain products or producers in terms of
sustainability. Collaboratively, they determine the color in an open and transparent
way. There should be an online platform, a "Github of the economy", where the color
determination takes place and is documented. Every interested individual can view
how the decision on color was reached and can make suggestions, even though only
few have the required experience and the time to actually make decisions.


## More possibilities

If the community wants to, it can incorporate more into the colors than simply
social and environmental sustainability. For instance, few large corporations
usually create a distorted market situation. Large corporations often have
advantages due to their bargaining power and "economy of scale" effects. However,
when these advantages lead to a monopoly or an oligopoly, there is no fair price
determination any more. Consequences can be too high prices and exploitation.

If one wants to avoid oligopolies and create a more diverse market with many
small or medium sized players, then one could add a component to the color that
considers the company size (or rather its "market power"). With that, the
competetive advantage of large corporations could be reduced or removed so that
small companies might enter the market more easily and be more competetive
despite of their small size.



## The hard problem of finding color

Determining a fair color for virtually every product or service that can be bought
seems like such a colossal task that it might well be deemed impossible. It might
mean performing audits similar to what is done currently for products to obtain
certain labels on sustainability or fair trade, but not for some specific products,
but for everything. On the other hand, there already exist software tools and databases
for performing lifecycle analyses (LCA), which could be leveraged to estimate products'
environmental impact, an important part of the color determination. Also, if products
consist of components, then the color can partly be a composite. If the same
components are used in other products, one can re-use their color.

But still, several problems come to mind: One is the question how the transition
to the new currency might be done, a classical hen-egg-problem. If the currency
is not in wide use, only very few products will have an adequate color. This means
only few products can be bought with the new currency in a meaningful way, which
might attract few adopters, leading to very limited usage.

Another problem is that of scalability: while it is in principle possible to
determine color, it is very hard to do this completely, including every aspect,
and for every product. Of course, perfection is impossible to reach, but one
should at least take a closer look, which takes time. A lot of manpower is
required to do the audits, which might need to include travels and on-site
visits. Some might argue that the work of the auditors would be so expensive
that it would be hard for society to pay for it. On the other hand, it should be
considered if an ethical and fair (and thus socially stable) economy is worth the
price. Thinking in degrowth terms, people in post-industrial economies ought to
work only half as much, both because the economy should be less "productive" and
because it would free a lot of "off-market" time. This freed time could be partly
used for the color determination. But still, any idea to simplify the color
determination process and save (human) resources as much as possible would be
very welcome.

A third problem is that of changing time: The color of a product can never be
constant, but must continuously adapt and react to changes in the production.
This is important: to reward producers when they improve, to create good
incentives. But also to prevent fraud where producers change the production
shortly after the audit. This adds to the colossal nature of the endeavour
and makes it a kind of "Sisyphos' work".

A fourth problem is that one must make sure that the system is not corrupted or
gamed. Of course, there will be strong motivation to come out "as green as
possible", so producers will "do their best", both on legal and illegal paths.
The transparency of making all color decisions public on a Github-like platform
might help, but probably will not be enough. It is clear that the color determiners
would be in an easily corruptible position and it might be a good idea to reward
them well to make them more independent from "gifts" from the industry.


<!-- Even though not completely secure from corruption and manipulation, the open nature of the color determination -->
<!-- might force stakeholders to accept a bad color, because an interference would be -->
<!-- visible to the public. Another aspect would be to show the color of a purchase -->
<!-- to each consumer before payment (or already on the price tag), therefore allowing -->
<!-- consumers to buy more consciously, apart from the perhaps existing price advantage. -->

### Summary of open questions

To summarize the afore-mentioned open questions:

1. *Hen-egg:* How to transition the economy to the use of the new currency?
   How to attract consumers? How to attract producers?
2. *Scalability:* How to determine color for "everything"? (Too much work?)
3. *Updates:* How to keep track of changes in the real world?
4. *Corruption:* How to prevent corruption and misuse?

<!-- 5. Who decides about the overall up and down valuation? -->
<!-- 6. Who decides what the "red money fund" is used for? (What external costs -->
<!--    should be alleviated?) -->

This is not meant to be a complete list. There may be many additional issues.
For instance, a whole category of open questions revolves around the democratic,
decentralized organization of the network. One must find ways to efficiently
make collective decisions, without introducing centralized structures and
hierarchy that would involve a lot of trust and opportunity for corruption.
An interesting idea might be the concept of "liquid democracy" [@Ford2002; @Ford2014; @Bauwens2014].

There might be many technical hurdles as well, for example energy-efficient,
secure, distributed validation of the blockchain(s).


## Attracting a critical mass

If the general public is ever to embrace this new technology, it is immensely
important to make it attractive not only to a few people hat are both tech-savvy
and care about the planet.

One strategy could be to create a subculture that identifies with this new technology
as part of their lifestyle. To make the connection stronger, one could make use
of good design, e.g. a very smart name, a fitting logo and a "cool"-looking user
interface. The subculture embracing the new technology could cause the public to
talk about this new phenomenon, raising more and more interest. Eventually, the
subculture may blend into the mainstream and become part of everyone's lives.

However, more important than appealing design will be usability: ease of use
on one hand (paying with the app must be as or even more intuitive than writing
a text message), and a perceived gain on the other hand. If the new technology
can buy you nothing you can buy with your trusted currency already, and if prices
are even higher, then the average customer will not use it, no matter how
"sustainable" it is. There must be some advantages to using it. For example,
if a new economy would be built around it, there might be products (perhaps
highly sought-after innovations) that are only sold in the new currency.

Another interesting strategy would be to create a parallel economy for the new
currency that has as many closed circles as possible. This is the concept followed
by most complementary currencies. Apart from the new currency enabling access
to this somewhat "privileged club", its use might also have economic advantages.
If the alternative currency has more value inside these closed circles than the
conventional currency of the outside world, then people will choose to use it.
Especially in times of financial crises troubling conventional currencies. This
is what happened in some experiments like the Wörgl experiment [@wiki:Wörgl] or
the Wära experiment [@wiki:Wära], for instance.

It is obvious that it is very important to find early adopters not only among
buyers, but even more so among sellers: i.e. producers and merchants (and corresponding
auditors). Only if one can buy things, buyers can start using it. Naturally, small
companies that already today care a lot about sustainability might be the first
adopters. Conservative companies that would be forced to change their behavior
will only join when they absolutely have to if they want to survive.

The CoinSence community (see http://coinsence.org/) is similar to the Viridian
project and aims to establish complementary social currencies. Within the CoinSence
community, there already exist more developed plans on how to attract people and
businesses and how to achieve adoption. There might be opportunities to cooperate
with CoinSence or to incorporate Viridian into CoinSence.

### Challenging the *homo oeconomicus*

On a side note, there are more motivations for people's actions than selfish advantage.
Many surveys have shown that, when being asked openly, a vast majority of
individuals agrees to support social or environmental issues [see @Klein2014, p. 118; @Eurobarometer].
This is because everyone wants to appear as a "moral", "good", and "social" person towards others.
One might call this a kind of "social peer pressure". But this social peer pressure
is completely removed from economic activity. Instead, the neoclassical view of
economics has taught us that it is a law of nature that every human must act "selfish"
in the economy [see @AdamSmith, Book I, Chapter II, p. 32]. This leads, in the
absence of peer pressure, to the famous "tragedy of the commmons" [@wiki:Tragedy_of_the_commons].
However, communication between individuals, applying peer pressure, can solve
this tragedy and it works for many commons across the world [@NobelOstrom].

Thus, if some kind of peer pressure would be incorporated into economic transactions
by means of cryptocurrency algorithms, it might also help to make economic decisions
more balanced and less one-dimensional.
The human brain consists of different parts, some of which are responsible for
rational, others for emotional behavior. One idea behind the colorization of money
is that the process of spending money stimulates the rational side of the brain,
so that emotional aspects such as empathy are not considered. However, when people
have the task of determining the color of a certain good---without being in the
situation of buying it---aspects such as the effect on other people
and on the environment will be more important and the emotional side of the brain
will be stimulated as well. This might offer a counterbalance to the
consumerism of today.[^yinyang]
Thus, the colorization of money transactions might be seen as an application
of social peer pressure (via the color determining community) to trade. 

<!-- Who would openly deny a societal consensus that takes into account
social and environmental values, leading to benefits for everyone? -->
If not using the colored currency would appear immoral and misanthropic, then
there might be a strong socio-psychological incentive to use it.
The psychological effect might be even stronger when showing the transaction color
to the customer before the transaction is carried out.

[^yinyang]: Bernard Lietaer [@Lietaer2012] refers to the duality of *yin* and *yang*
from Chinese philosophy and speaks of money being almost entirely associated with
*yang* attributes like aggressiveness, strength, competition (considered dominantly
male) and not with *yin* attributes like care, empathy, softness (considered dominantly
female). Lietaer perceives a lack of appreciation of *yin* values in modern society,
caused by the dominance of *yang* money. In this sense, colorized money might add a
missing *yin* component. XXX Look up if this is mentioned in Lietaer2012


# Possible implementation

From a software architecture point of view, there need to be different components
to make the currency system work:

1. A distributed database to store the colors that can be queried during
   transactions.
2. A collaborative platform, from which the color database can be edited,
   perhaps similar to Github.
3. The cryptocurrency itself (a distributed ledger storing the transactions),
   for example as a blockchain.
4. Applications for making transactions and interact with the cryptocurrency.


## The color database or "color blockchain"

The color database gives a unique ID to each product. It is updated frequently
and always returns the current color when queried. It may or may not be a blockchain.
A blockchain would have added benefits like tamper resistance and dectralization
as well as version control and traceability. If it is a blockchain, then there
would be an "account" for every product and the acount's balance would be the
current color value, which can be changed with "color transactions" by the color
determiners, with added meta-data on why the color is changed. Perhaps, it makes
sense to store not only the resulting color itself, but also "subcolors" for
relevant aspects: social sustainability, environmental sustainability, longevity
of the product, size (or "market power") of the company, etc. Maybe the color
database can incorporate user ratings as well, e.g. rating the usefulness or the
quality of the product in use.
<!-- XXX Or would the color be part of the cryptocurrency itself, so not an oracle -->
<!-- for a smart contract, but be stored in the transaction blockchain as well? This -->
<!-- way, the currency's blockchain would also provide supply chain tracking, at least -->
<!-- in terms of sustainability. But the blockchain in a traditional sense can only -->
<!-- store transactions that can happen only once, not "products" that can be sold -->
<!-- n times. So, in the blockchain, one should store each transaction's color, -->
<!-- a timestamp, and an ID of the product, but the current product ID's color must -->
<!-- come from an external source. (Or not?) -->
<!-- XXX Or we need two separate blockchains: One for transactions (like bitcoin), one -->
<!-- for colors. In the color blockchain, each product has an "account" and the account's -->
<!-- balance is its color that can change over time. -->

The separate color blockchain would be managed by the individuals determining colors,
the "color miners". Color mining is a task quite different from the transaction
mining, which secures the transaction blockchain. Color mining involves both auditing
or LCA work to gather information backing the color value, but also peer reviewing
other miners' "color transactions" before they become a legitimate part of the
blockchain. Each color miner registers an account on the transaction blockchain, to
which rewards for their work are transferred.


## The collaborative platform

It must be open to anyone. For reading the platform's content, no registration is
necessary. Anyone registering on the platform is becoming a "color miner"
who helps in the determination of colors. Like there can be "projects" in Github,
there are workspaces for each product and service.

One has to make sure that there is only one "genuine" product workspace and that
changes to the color of a product are only done by trusted members. For this
purpose, one could introduce a user permission system like a "web of trust",
where only high rated members have permission to create new work spaces and assign
write permissions of certain workspaces to certain members. The downside is that
this makes the system more complicated, increases the threshold to add new products
and introduces hierarchy. Any better ideas are appreciated.

An interesting idea for the decisions on colors is the use of liquid democracy
tools, which represent "a fast, decentralized, collaborative question-answering
system, which works by enabling chained answer recommendation" [@Bauwens2014].
This would allow a broader group of people to do informed votings on colors,
even though only a small group of experts provides the relevant background
information (e.g. from audits and visits). When a broader group decides
democratically about color, it makes abuse more difficult. Another idea to prevent
abuse is to establish peer review within the group of color miners. For example,
one could establish that each new information or decision about color has to be
approved by at least two other randomly selected independent color miners.

The CoinSence community (see http://coinsence.org/) has developed a working
online platform for information exchange, collaboration and the creation and
exchange of complementary (social) currencies. They plan to incorporate
liquid democracy tools into their community platform (https://community.coinsence.org),
so it might be viable to build the color determination platform on top of
the CoinSence platform.


## The cryptocurrency or "transaction blockchain"

For illustrative purposes and for a more "hands-on" feeling, I will draft a
minimal implementation of the cryptocurrency in the Ethereum framework. Keep
in mind, however, that the properties of the Ethereum blockchain may not be
suitable for the endeavour.

Ethereum is a programming framework for smart contracts. Smart contracts are a
generalization of cryptocurrencies, where a blockchain can be used for different
purposes. Ethereum is completely independent of the most prominent cryptocurrency,
bitcoin, and has its own blockchain and its own cryptocurrency called *ether*,
which is not really used as a currency, but merely as a means of carrying out smart contracts.
With smart contracts, computer code can be attached to the blockchain, which is
executed by the miners when the transaction is validated. Therefore, it can be
made sure that the content of a contract is carried out as intended because the
code is stored, secure from manipulation and distributed, in the blockchain.

One of the simplest things that can be implemented with Ethereum is a new cryptocurrency.
Ethereum features an own programming language for smart contracts (a bit similar to
JavaScript): Solidity. Taking a sort of *Hello World!* example of an Ethereum
smart contract [@etherdoc] and changing it slightly, one arrives at a first
rather trivial implementation of the new cryptocurrency:

```javascript
contract Coin {
    address public minter;
    mapping (address => uint) public balances;
    Product public product;

    event Sent(address from, address to, uint amount);

    function Coin() public {
        minter = msg.sender;
        product = Product(msg.prod_id);
    }

    function mint(address receiver, uint amount) public {
        if (msg.sender != minter) return;
        balances[receiver] += amount;
    }

    function send(address receiver, uint amount) public {
        if (balances[msg.sender] < amount) return;
        balances[msg.sender] -= amount;
        c = product.getColor();
        balances[receiver] += colorize(amount, c);
        Sent(msg.sender, receiver, amount);
    }
    
    function colorize(uint amount, double c) public {
        return uint(amount * (1. + c));
    }
}

contract Product {
    address public id;
    
    function Product(address prod_id) public {
        id = prod_id;
    }
    
    function getColor() public {
        ... // query color DB
    }
}
```

The application of the product's color is capsuled in the function `colorize()`.
This is the place where the used algorithm, here simply $y = (1 + c) \; p$, would
be adjusted.

The key function is the still unimplemented function `getColor()` in the `Product`
contract. The color of the transaction is an information that must be fetched
from the outside non-digital world. In smart contract jargon, a provider of such
"external information" is called an *oracle*, e.g. see http://www.oraclize.it/.
It is such an oracle that needs to be built in form of the collaborative platform
where the colors are determined, stored in the distributed color database, which
would be queried by the function `getColor()` to fetch the currently valid color
value of the product.

### Blockchain validation or "mining"

With the blockchain being a *distributed* ledger, there is no central authority
to decide about the validity of transactions. This leads to the famous
"double-spend" problem of cryptocurrencies. How to prevent an attacker from
spending an amount of money multiple times? When a bank supervises transactions,
it is in charge of the only central ledger and can allow the first transaction
(emptying the bank account), but will cancel the second transaction (because the
account is already empty). With a distributed ledger, there is one copy of the
ledger on each node participating in the distributed network. In principle, one
could send the first transaction to one node and then, before the first node has
had a chance to spread the information, send a second transaction to another node.
The second node will not know that the account was already emptied in the first
transaction. Therefore, there must be a process of consolidation between all nodes.

<h4>Proof-of-work</h4>

The ultimate question is: Who is to decide about what set of transactions is
accepted (and added to the blockchain)? Because we don't want to trust a central
authority, this power has to be spread across the network. Usually, it passes
from node to node for each new block, where a block is a small part of the blockchain
comprising a certain number of transactions. There are different algorithms to
decide about the next node whose turn it is to validate a block of transactions.
In Bitcoin [@Nakamoto] and some other cryptocurrencies, this algorithm is called
*proof-of-work*. It involves computing a number ("work"), which must have a
certain property, by random trial and error. The more computing power one has,
the sooner one will (on average) find a suitable number by chance. The fraction
of transactions one can validate is proprtional to the owned fraction of
computing power of the total network. The result of this is that, given many
participants, it is extremely resource-intensive and therefore expensive to gain
control over the network. One would need to contribute more than 50% of the entire
computing power to take over the network and make it accept fraudulent
transactions.[^bitcoin_detail] In the end,  The elegance of this is that Bitcoin stays honest
and secure without the need to trust anyone.

[^bitcoin_detail]: To understand this, one needs to know that the nodes can
disagree on the blockchain validation. If a dishonest node validates a block of 
transactions containing a fraud (e.g. a double-spend), then the other honest nodes
will not accept that block and will create an alternate version of the blockchain
where the fraudulent transaction is removed. The blockchain splits into two
versions. This is called a *fork*. The nodes will choose the blockchain version
that is longer, because it means more nodes have trust in this version. The
dishonest attacker needs to add more blocks to the chain than all the honest
nodes together, therefore more than 50% of the computing power is needed.

One downside is that it is not only extremely resource-intensive to take over
the network, it is also extremely resource-intensive to run it in the first
place. Because the block validating nodes are rewarded with new bitcoins, they
are also called "miners" (they mine for digital gold). The proof-of-work system
incentivizes the installation of a lot of computing power because the more
computing power one owns, the more often one will validate a block and obtain
the reward. The result is an energy consumption as high as an entire country's
[@BitcoinEnergy], not for computing anything valuable, but only for winning the
right to validate transactions. Another downside is that the system increases
inequality: the rich can afford to install a lot of computing power, by which
they get even richer because they obtain a large share of the reward. Because of
that, there is even a dangerous tendency towards mining monopolies or oligopolies,
where big mining farms buy smaller competitors until only few (or one) very large
players are left.

<h4>Proof-of-stake</h4>

A variation of the proof-of-work algorithm is called *proof-of-stake* [@wiki:Proof-of-stake],
in which the expensive computation is omitted. The creator of the next block is
selected randomly, however favoring nodes that have higher "stakes", i.e. own more
of the cryptocurrency. Also, the "age" of the coins can be considered,
not just the amount of coins. The idea behind this is that nodes that hold higher
stakes (for a longer time) have contributed longer and are thus more trustworthy
than recently joined nodes. Another advantage of proof-of-stake is that the owners
of the currency are the ones that have the right to secure it. In proof-of-work,
the miners might sell their entire rewards for another currency and thus have
little incentive to use the complementary currency and invest in its ecosystem.

A disadvantage is that, similar to proof-of-work, rich participants are preferred
so that again inequality tends to increase. Most cryptocurrencies that use proof-of-stake,
e.g. BlackCoin [@Vasin] and Peercoin [@King2012], use mechanisms to prevent too
much domination from single rich members, however. The Ethereum blockchain also
uses proof-of-stake.

<h4>Proof-of-whatever?</h4>

It is important to come up with a well-designed validation algorithm that minimizes
energy consumption as much as possible (for sustainability), that prevents
inequality, oligopolies and centralization, and that is reasonably secure (i.e.
not too much prone to fraud).

More ideas for improved alternatives to proof-of-work and proof-of-stake exist
already. A promising approach may be the *proof-of-cooperation* [@König2016;
@König2017] of FairCoin, a cryptocurrency developed by the FairCoop cooperative.
Both proof-of-work and proof-of-stake rely on competition between miners.
Proof-of-cooperation, however, relies on the assumption that cooperation is more
efficient than competition. It works roughly like this: Block creation is evenly
shared between nodes. Every node decides independently about the node whose
last block creation lies furthest back in the past (by looking at the blockchain
history). The node sends a
<!-- cryptographically signed -->
message with its decision over the network. When an assigned node has received
enough messages from the other nodes approving it, it goes ahead and creates the
new block. This mehcanism could still lead to centralization when a single
entity participates with a very large number of nodes. This is currently prevented
in FairCoin by validating each node (by talking to them and perhaps even visit
them in real life) before it goes online. This validation, as well as decisions
about blockchain management, are performed by a single central organization, the
FairCoop assembly and administrators appointed by the assembly, which means that
it is not truly decentralized and might have a vulnerable point that may be
corruptible by very powerful entities. I also suspect that the FairCoin
infrastructure does not scale well. FairCoin has a goal of operating 40--50
validated nodes, with a hardcoded maximum of 100 nodes, mostly running on
low-power Raspberry Pis. This can hardly be suitable for running the entire
world economy. But still, it's a very interesting concept that may be a good
starting point.

There are also other ideas, even beyond the blockchain. For example, there are
cryptocurrencies that use *directed acyclic graphs* (DAGs), a sort of
generalization of the blockchain. The most prominent are currently IOTA [@wiki:IOTA;
Popov2017] (https://iota.org/) and Hashgraph [@Baird2016] (https://hashgraph.com/),
however, the latter is patented, which poses a problem for accessibility and wide-spread
adoption. IOTA comes from the IOT community as a zero-cost micro-payments
cryptocurrency. It claims to be much more scalable than a Bitcoin-like blockchain,
with transaction validation becoming faster with the number of transactions, not
slower like in Bitcoin. IOTA also claims to be, like FairCoin, cooperative instead
of competetive [@Semko2017].


## Interfaces to the "transaction blockchain"

Of course, aside of the blockchain backend, there must be frontend applications
("wallets") for interacting with the blockchain, i.e. for making transactions
and paying things. There should also be very easy to use mobile applications
available for all smartphone operating systems, in addition to platform-independent
computer software.

In addition to conventional crypto-wallet features, there can be features
enhancing the information and awareness of the users. The color of the
transaction can be made visible, along with information on different components
(social, ecological, etc.) of the color decision and maybe even supporting
information leading to the decision and the history of the color value.


# Conclusion

Leveraging the wisdom of crowds and the power of cryptocurrencies, perhaps one
can reach a societal consensus towards a world that is more worth living in for
everyone, preserving life for future generations, while not giving up modern
lifestyles.

These are the requirements that must be fulfilled by the alternative currency:

1. It must feature a significantly more sustainable and/or fair economy.
2. It must be based on societal consensus to be accepted by (almost) everyone.
3. It must be very easy to use.
4. It must be fun to use and/or provide a real advantage.
5. It must be relatively hard to break the rules or abuse the system.

I assume that (1) would be a consequence of all the other points fulfilled, if
the currency would be designed similar to what is described above. (2) is the
main problem remaining to be solved: how to determine the external costs of goods
in a socially acceptable way? The idea to use crowdsourcing was only vaguely
described here. More thoughts on this would be greatly appreciated. (5) would only
be partly fulfilled by the use of cryptocurrencies, how to secure the crowdsourcing
against abuse is another problem to be solved. (3) and (4) have only been merely
scratched at the surface.

Therefore, we are still at the very beginning, but I hope that something good will evolve out of this.

# References