---
title: "The Viridian Project: A digital currency to internalize external costs for a more sustainable economy"
author: "Markus Voge"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: elsevier-with-titles-alphabetical.csl
# http://editor.citationstyles.org/searchByName/
output: html_document
---

# Introduction

Among the biggest problems of humanity in the 21st century areenvironmental crises
such as global warming, but also social crises such as inequal distribution of wealth,
with some still suffering from hunger. These problems in turn are catalysts for conflicts
and migration, putting entire societies under stress. Many experts say that we are
pushing the Earth's systems beyond their planetary boundaries [@Rockstrm2009; @Barnosky2012],
and that the current lifestyle of the Global North would require more than one planet to
survive in the long term.[^overshoot]

[^overshoot]: The Global Footprint Network annually calculates the date of the
*Earth Overshoot Day*, on which humanity has depleted the amount of natural resources
that is regenerated each year, see e.g. the 2016 annual report [@GFN2016],
https://www.footprintnetwork.org/our-work/earth-overshoot-day/, or https://www.overshootday.org/.
In 2017, the Earth Overshoot Day fell on 2 August.

Most of these problems have to do with economic activity and the underlying monetary
systems [@Lietaer2012]. One well-known problem is the privatization of profits, while
costs, e.g. of  environmental damages, are being socialized [@wiki:Externality; @wiki:Lemon_socialism].
As a result, often small elites are becoming wealthier, while a majority's well-being is
stagnating or even declining. Another result is the acceptance (also by the suffering majority)
of environmental damage as an inevitable consequence of economic activity and progress,
even when it is risking to destroy the very foundation of life and when healthier
alternatives are both available and financeable by society [see @Klein2014 pages 12ff., 113ff., 117ff.].
One problem of established monetary systems is the inevitability of unlimited economic
growth, which is, as many have pointed out [@Meadows1974; @Georgescu-Roegen1971; @Schneider2010; @Hopkins2014],
opposed to the fact of living on a planet with limited resources. The economic growth
is largely driven by the financial sector and the habit of paying interest rate and
return of investment: companies acquiring money are forced to grow to reel in more
than they invested and pay back their donors. For instance, this drives fossil fuel
companies to ever riskier extraction practices, such as fracking, deep sea drilling,
or drilling in the Arctic, with sometimes catastrophic consequences [see @Klein2014 pages 146--48, 332f., 144].

# Changing the monetary system

As turned out by Belgian economist Bernard Lietaer [@Lietaer2012], all of these
problems might be addressed by monetary innovations. One interesting idea is the
concept of complementary currencies: Currencies that have limited validity, e.g.
only in a certain geographic area or only within certain branches of economy.
These currencies are not to replace national or international currencies, but to
complement them and thereby strengthen closed circles and reduce extreme aggregation
of wealth. The idea of complementary currencies is already actively followed by
people from several groups, e.g. the Transition Towns movement [@Hopkins2014] that
aims to build a post-fossil, sustainable, and resilient economy and by the degrowth
movement [@Douthwaite2012; @Dittmer2013; @Hornborg2017] that aims to put an end to
unlimited economic growth. Complementary currencies are already in use at some places
and prove to strengthen regional and more sustainable value generation. In addition,
there are more ideas to redesign our monetary systems, e.g. *The Chicago Plan Revisited* [@wiki:The_Chicago_Plan_Revisited; @IMF2012],
a report issued by the International Monetary Fund (IMF) in 2012, suggesting that
private banks should only be allowed to lend money that they have in deposits
(100% reserve banking), reducing both public and private debt and resulting in a
more balanced and stable economy. Another idea is to establish a complementary
currency on the national level in each country, distributed to all citizens as a
basic income, thereby "insulat[ing] local sustainability and resilience from the
deleterious effects of globalization and financial speculation" [@Hornborg2017].

Another monetary innovation are cryptocurrencies (the most popular being Bitcoin)
based on a distributed ledger or blockchain, which do not require any kind of central
institution like a bank [@Nakamoto]. Besides the benefit of reduced cost, a much
more fascinating aspect of cryptocurrencies is that of the democratization of money.
Instead of restricting the privileges of the creation of new money and the verification
of money transactions to few centralized (mostly private) banks, giving them a great
deal of power, this power is distributed by giving it to a large group of relatively
equal players, which everyone can join. A downside is that cryptocurrencies can only
exist digitally and thus require the use of electronic devices. This creates a vulnerable
dependency on electricity and might increase resource consumption,[^crypto_consumption]
which is not good from a sustainability perspective. On the other hand, smartphone
devices capable of digital payments are already now almost ubiquitous, even in developing
parts of the world.

[^crypto_consumption]: In addition to increased use of electronic devices for performing
payments using cyrpto-wallet apps, an enormous energy and resource consumption can be
generated when transactions are verified in a cryptocurrency's network using
so-called *proof-of-work* computations. The Bitcoin protocol is based on proof-of-work
and the Bitcoin network's global energy consumption is assumed to exceed that
of many European countries, including Ireland [@BitcoinEnergy]. There are some
ideas on how to replace proof-of-work with less resource-intensive concepts, e.g.
the *proof-of-stake* used by the Ethereum blockchain.

Cryptocurrencies are not controlled by central institutions, banks and/or governments,
i.e. individual people, but by mathematic algorithms carried out by computer software,
distributed on all participants' devices. This offers several opportunities:
In principle, these algorithms can be formulated in any imaginable way, enabling
the easy implementation of a very different monetary system. Also, if designed properly,
it might be much harder to break the rules of the currency system, since computers,
unlike people, are not corruptible. Especially if manipulation on a majority of
devices would be necessary, then this is virtually impossible, which is also what
makes Bitcoin a very secure currency.


## A new ethical cryptocurrency

The idea is to combine the aim of a more sustainable economy with the benefits of
cryptocurrencies. If cryptocurrencies can be designed as one sees fit, then it might
be possible to solve the problems of conventional monetary systems.

As a starting point, one can try to incorporate external costs into prices of
goods and services. External costs are those that are not part of the economic
calculation when a price is determined, however they still arise (sometimes at a
different place and/or time, harming third parties) and have to be paid by someone,
usually the tax payer or residents of affected areas. Examples of external costs
are health damages caused by pollution or costs of disasters that are the result
of careless extraction of resources, including more frequent extreme weather events
as a result of green house gas emissions. In order to ensure that external costs
are taken care of as well as discourage harmful and encourage sustainable behavior,
it is better to internalize external costs into the actual price.


## Two-dimensional money

The problem is that a (traditional) currency system cannot directly influence
the way a business owner determines prices. The currency as a whole can be valued
up or down by increasing or reducing the amount of fresh money (i.e. setting the
key interest rate), but that does not influence individual trades. However,
cryptocurrencies basically consist of transactions between peers and may offer
a possibility to interfere right there. A possible idea is to add another dimension
to money.

Ever since the invention of money, it has been one-dimensional in its very nature,
the single dimension being value, or rather quantity. Money is a single scale,
against which almost every thing can be measured. Digital currencies might remove
this restriction and could add other dimensions to quantity, for example quality.
Not only the mere amount of (physical) wealth would matter, but also the change
in quality of life that comes along with the aggregation of wealth.

To make it more visual, one can think of the new dimension as color:[^colored_coins]
From green, indicating good quality, to red, indicating bad quality. For means of
computation, color can be converted to a number between -1 (red) and +1 (green).
Since it would be complicated to handle real 2D money (e.g. adding or subtracting
money involves vector calculus, which is simple, but not convenient for many people),
the color should only come into play during a transaction, on the receiver's side.
When the money enters a wallet, it is automatically projected onto a single axis.
This means that red money is valued down, while green money is valued up, which
corresponds to a diagonal movement in the 2D money space.

XXX Add Figure to visualize color of money (2 axes: y axis is value, x axis is color)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(fields)
library(grImport)
library(igraph)

# https://stackoverflow.com/questions/28042220/how-to-make-custom-plot-symbols-from-vector-graphics-in-r
PostScriptTrace("coin.eps") # creates .xml in the working directory
coin <- readPicture("coin.eps.xml")

y <- c(0, 2000, 1500, 300)
x <- rep(0, length(y))
plot(x, y, bty="n", xlab = "Color c (Quality)",
     ylab = "Quantity: $/€/£/¥/...", ylim = (max(y) - min(y)) * c(-0.187, 1.119))
xx = grconvertX(x = x, from = 'user', to = 'ndc')
yy = grconvertY(y = y, from = 'user', to = 'ndc')
grid.symbols(coin, x = xx, y = yy, size = 0.05)
# par(new=TRUE)
# plot(c(-1, 1), c(0, 0), axes = FALSE, xlab = "", ylab = "")
colorbar.plot(0, -0.149 * (max(y) - min(y)), seq(-1, 1, 0.01), strip.length = 1.7,
              col = colorRampPalette(c("#009900", "#990000"))(256))
# https://kbroman.wordpress.com/2012/10/11/curved-arrows-in-r/
invisible(igraph:::igraph.Arrows(0, y[1], 0, y[2], h.lwd=2, sh.lwd=2,
                                 curve=-0.3e-3, width=1, size=0.7))
invisible(igraph:::igraph.Arrows(0, y[2], 0, y[3], h.lwd=2, sh.lwd=2,
                       curve=0, width=1, size=0.7))
invisible(igraph:::igraph.Arrows(0, y[3], 0, y[4], h.lwd=2, sh.lwd=2,
                       curve=0, width=1, size=0.7))
# https://stackoverflow.com/questions/30658177/r-plot-with-axes-in-the-center
# arrows(-1, 0, 1, 0, lwd=1, length=0.15)
# arrows(0, 0, 0, 15, lwd=1, length=0.15)
```

XXX Add Figure to visualize money revaluation by projection on y axis

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(fields)
library(grImport)
library(igraph)

# https://stackoverflow.com/questions/28042220/how-to-make-custom-plot-symbols-from-vector-graphics-in-r
PostScriptTrace("coin.eps") # creates .xml in the working directory
coin <- readPicture("coin.eps.xml")

y <- c(0, 2000, 1500, 300)
x <- rep(0, length(y))
plot(x, y, bty="n", xlab = "Color c (Quality)",
     ylab = "Quantity: $/€/£/¥/...", ylim = (max(y) - min(y)) * c(-0.187, 1.119))
xx = grconvertX(x = x, from = 'user', to = 'ndc')
yy = grconvertY(y = y, from = 'user', to = 'ndc')
grid.symbols(coin, x = xx, y = yy, size = 0.05)
# par(new=TRUE)
# plot(c(-1, 1), c(0, 0), axes = FALSE, xlab = "", ylab = "")
colorbar.plot(0, -0.149 * (max(y) - min(y)), seq(-1, 1, 0.01), strip.length = 1.7,
              col = colorRampPalette(c("#009900", "#990000"))(256))
invisible(igraph:::igraph.Arrows(0, y[1], 0, y[2], h.lwd=2, sh.lwd=2,
                                 curve=-0.3e-3, width=1, size=0.7))
invisible(igraph:::igraph.Arrows(0, y[2], 0, y[3], h.lwd=2, sh.lwd=2,
                       curve=0, width=1, size=0.7))
invisible(igraph:::igraph.Arrows(0, y[3], 0, y[4], h.lwd=2, sh.lwd=2,
                       curve=0, width=1, size=0.7))
# https://stackoverflow.com/questions/30658177/r-plot-with-axes-in-the-center
# arrows(-1, 0, 1, 0, lwd=1, length=0.15)
# arrows(0, 0, 0, 15, lwd=1, length=0.15)
```

<div>Icons made by <a href="https://www.flaticon.com/authors/smashicons" title="Smashicons">Smashicons</a> from <a href="https://www.flaticon.com/" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div>

[^colored_coins]: Within the cryptocurrency community, already since 2013 there
exists the term "colored coins" (see e.g. http://coloredcoins.org/, https://www.coinprism.com/),
which has nothing to do with the concept described here. "Colored coins" are
bitcoins carrying certain information. They can be used for example to prove
ownership of a certain asset. It is called "colored" because a bitcoin is
"marked" to make it distinguishable from all other bitcoins. Since the ownership
is linked to a cryptocurrency coin, it can be easily transferred to another acount.
This enables the realization of "smart contracts", where digital or real world
assets can be sold securely using a blockchain. Colored coins are similar to
Ethereum, however not using a seperate blockchain, but the bitcoin blockchain.

It can be argued how exactly this projection should be done, e.g. in a linear or
non-linear way. It is also debatable if there should be upper and lower limits to
the de-/revaluation and what these limits should be. To avoid too extreme value
changes, the effect of the color on the transaction should be considered when the
color is determined. For now, a very simplistic choice is made: A color $c$ of a
transaction changes the value of the reserved money by factor $1 + c$, which can
lie anywhere between 0 (dark red, complete devaluation, i.e. selling is prohibited)
and 2 (dark green, i.e. duplication of value). This means that there is also
"neutral money", which is yellow, $c = 0$, that does not change value, $1 + c = 1$.
So, the price $p$ paid by the buyer is changed into the seller's yield $y$ via:

$$ y = (1 + c) \; p $$

It is also an interesting question whether there should be more up or more down
valuation, which would essentially either create or destroy money, or whether they
should balance each other. This could be controlled via a regularly changing overall
normalization of the re-evaluation, so that the average change in the total amount of
money is constant. Currently in use mainstream currencies constantly create new
money via credits, which creates a low inflation rate. Bitcoin on the other hand,
the most prominent cryptocurrency, has a fixed volume of 21 million bitcoins.
This limited amount of money leads to a deflationary currency. Mainstream economists
argue that it is beneficial to have a low inflation rate, because it prevents people
hoarding money and creates an incentive to spend the money and bring it back into
circulation. This might lead to a more equal distribution of wealth. On the other hand,
it might lead to unnecessary consumption of goods and thus be a driver to (excess)
economic growth. Degrowth economists therefore might favor a deflationary currency.
There are claims that interest rate (and therefore inflation) incentivizes short-term
investments, while an interest-free currency system with demurrage (money losing
some if its value over time, i.e. negative interest rate, or money "disappearing")
incentivizes long-term thinking and promotes a more sustainable economy [@Lietaer2012].

### An example

As a concrete example for a transaction in the new currency, buying an emission-intensive
car with a fossil fuel burning engine might involve red money, while an emission-free
electric vehicle is bought with green money. 

Let's assume that the seller of the red money car had to pay 9,000 coins to buy
the car from the producer. The car seller needs a profit margin of 1,000 coins,
so would sell it for 10,000 coins without considering color. The color of the
transaction is -0.5, meaning that the yield $y$ will come out 50% lower than the
price $p$. For $y$ to be 10,000, $p$ must be $p = y / (1 + c)$, so the car must
be sold for 20,000 coins.

On the other hand, the green car seller might have paid 15,000 coins to the
producer and wants to earn 1,000 coins as well. The color is 0.1, translating
into a price of $16,000/1.1 = 14,545.45$ coins, making the car cheaper for the
consumer even though it was more costly to produce.

Upon entering the seller's wallet, the red money loses some of its value, while
the green money is valued up. This effectively means that the fossil fuel car will
be sold with a higher price (since the seller gains less), while the electric car
gets cheaper. It is like a higher tax on one, while subsidizing the other. The
buyer does not even need to know about the color of the transaction or the process
of (de-)valuation, only the seller needs to include the color into their price
calculation. This makes it easier to adopt for the consumer. This can help cleaner,
emerging technologies, sometimes needed as quickly as possible from an ecological
point of view, to become marketable much quicker. It can create the right incentives
to make good investments and use good practices, while punishing behavior that is
harmful to society.

It is important to note that, similar to value added tax, the colorization should
probably not be applied multiple times, but only at the end consumer. Intermediate
reevaluations must be undone accordingly.

## Redistribution of value

The red money transactions may take a part of the buyer's money and pay it into
a fund to alleviate side-effect damages like environmental disasters or increased
health costs and can also fund dedicated projects supporting unemployed people
during structural change. This would adhere to the principle of "the polluter pays"
[see @Klein2014 pages 110--19] and would be considered fair by most people.

All transactions regardless of color should include a small transaction fee that
is paid to the color determiners to reward them for their services.

## Managed by the crowd

While a similar effect to that of colored money can be reached by taxation and
subsidization, the interesting aspect is that a cryptocurrency offers a possibility
to do this without a central authority involved. The main problem of transforming
the economy towards sustainable development is to do it in a way that is fair
and socially accepted by everyone. 

Past policies have failed to do so. The less wealthy will not accept a change
that feels like they have to give up their standard of living, while the wealthiest,
the ones most responsible, e.g. those profiting from extraction of resources, are not
paying their toll. The fear of losing jobs due to a necessary system change is
quickly sparked in the public and surely must be addressed, e.g. by dedicated projects.
But because governance fails to introduce such projects, which should be paid for
by polluters like the fossil fuel companies, system change never comes.

This is why such important decisions like the color of money cannot be placed
into the hand of a few privileged, or a single person. It must be done
collaboratively by the crowd, in a way that is acceptable by everyone.

This is an attemt to bypass government and any centralized institutions or elites:
Like the Bitcoin miners perform the service of validating transactions and keeping
the currency's network secure, dedicated decentralized groups could determine the
transaction colors. Much like developers of open source software, there could be
groups of people dedicated to auditing certain products or producers in terms of
sustainability. Collaboratively, they determine the color in an open and transparent
way. There should be an online platform, a "Github of the economy", where the color
determination takes place and is documented. Every interested individual can view
how the decision on color was reached and can make suggestions, even though only
few have the required experience and the time to actually make decisions.


## The hard problem of finding color

Determining a fair color for virtually every product or service that can be bought
seems like such a colossal task that it might well be deemed impossible. It might
mean performing audits similar to what is done currently for products to obtain
certain labels on sustainability or fair trade, but not for some specific products,
but for everything. On the other hand, there already exist software tools and databases
for performing lifecycle analyses (LCA), which could be leveraged to estimate products'
environmental impact, an important part of the color determination. Also, if products
consist of components, then the color can partly be a composite. If the same
components are used in other products, one can re-use their color.

But still, several problems come to mind: One is the question how the transition
to the new currency might be done, a classical hen-egg-problem. If the currency
is not in wide use, only very few products will have an adequate color. This means
only few products can be bought with the new currency in a meaningful way, which
might attract few adopters, leading to very limited usage.

Another problem is that of scalability: while it is in principle possible to
determine color, it is very hard to do this completely, including every aspect,
and for every product. Of course, perfection is impossible to reach, but one
should at least take a closer look, which takes time. A lot of manpower is
required to do the audits, which might need to include travels and on-site
visits. Some might argue that the work of the auditors would be so expensive
that it would be hard for society to pay for it. On the other hand, it should be
questioned if an ethical and fair economy is worth the price. However, any
idea to simplify the color determination process and save (human) resources as
much as possible would be very welcome.

A third problem is that of changing time: The color of a product can never be
constant, but must continuously adapt and react to changes in the production.
This is important, to reward producers when they improve, to create good
incentives. But also to prevent fraud where producers change the production
shortly after the audit.

A fourth problem is that one must make sure that the system is not corrupted or
gamed. Of course, there will be strong motivation to come out "as green as
possible", so producers will "do their best", both on legal and illegal paths.
The transparency of making all color decisions public on a Github-like platform
might help, but probably will not be enough. It is clear that the color determiners
would be in an easily corruptible position and it might be a good idea to reward
them well to make them more independent from "gifts" from the industry.

<!-- Even though not completely secure from corruption and manipulation, the open nature of the color determination -->
<!-- might force stakeholders to accept a bad color, because an interference would be -->
<!-- visible to the public. Another aspect would be to show the color of a purchase -->
<!-- to each consumer before payment (or already on the price tag), therefore allowing -->
<!-- consumers to buy more consciously, apart from the perhaps existing price advantage. -->


XXX Make actual enumeration of open questions
1. How to transition the economy to the use of the new currency? How to attract
   consumers? How to attract producers?
2. How to determine color for "everything"? (Too much work?)
3. How to keep track of changes in the real world?
4. How to prevent corruption?
5. Who decides about the overall up and down valuation?
6. Who decides what the "red money fund" is used for? (What external costs are alleviated?)


## Attracting a critical mass

If the general public is ever to embrace this new technology, it is immensely
important to make it attractive not only to a few people hat are both tech-savvy
and care about the planet.

One strategy could be to create a subculture that identifies with this new technology
as part of their lifestyle. To make the connection stronger, one could make use
of good design, e.g. a very smart name, a fitting logo and a "cool"-looking user
interface. The subculture embracing the new technology could cause the public to
talk about this new phenomenon, raising more and more interest. Eventually, the
subculture may blend into the mainstream and become part of everyone's lives.

However, more important than good-looking design will be usability: ease of use
on one hand (paying with the app must be as or even more intuitive than writing
a text message), and a perceived gain on the other hand. If the new technology
can buy you nothing you can buy with your trusted currency already, and if prices
are even higher, then the average customer will not use it, no matter how
"sustainable" it is. There must be some advantages to using it. For example,
if a new economy would be built around it, there might be products (perhaps
highly sought-after innovations) that are only sold in the new currency.

Another interesting strategy would be to create a parallel economy for the new
currency that has as many closed circles as possible. This is the concept followed
by most complementary currencies. Apart from the new currency enabling access
to this somewhat "privileged club", its use might also have economic advantages.
If the alternative currency has more value inside these closed circles than the
conventional currency of the outside world, then people will choose to use it.
Especially in times of financial crises troubling conventional currencies. This
is what happened in some experiments like the Wörgl experiment [@wiki:Wörgl] or
the Wära experiment [@wiki:Wära], for instance.

It is obvious that it is very important to find early adopters not only among
buyers, but even more so among sellers: i.e. producers and merchants (and corresponding
auditors). Only if one can buy things, buyers can start using it. Naturally, small
companies that already today care a lot about sustainability might be the first
adopters. Conservative companies that would be forced to change their behavior
will only join when they absolutely have to if they want to survive.

### Challenging the *homo oeconomicus*

On a side note, there are more motivations for people's actions than selfish advantage.
Many surveys have shown that, when being asked openly, a vast majority of
individuals agrees to support social or environmental issues [see @Klein2014, p. 118; @Eurobarometer].
This is because everyone wants to appear as a "moral", "good", and "social" person towards others.
One might call this a kind of "social peer pressure". But this social peer pressure
is completely removed from economic activity. Instead, the neoclassical view of
economics has taught us that it is a law of nature that every human must act "selfish"
in the economy [see @AdamSmith, Book I, Chapter II, p. 32]. This leads, in the
absence of peer pressure, to the famous "tragedy of the commmons" [@wiki:Tragedy_of_the_commons].
However, communication between individuals, applying peer pressure, can solve
this tragedy and it works for many commons across the world [@NobelOstrom].

Thus, if some kind of peer pressure would be incorporated into economic transactions
by means of cryptocurrency algorithms, it might also help to make economic decisions
more balanced and less one-dimensional. The colorization of money transactions might
be seen as an application of social peer pressure (via the color determining community)
to trade. Because who would openly deny a societal consensus that takes into account
social and environmental values, leading to benefits for everyone? If not using the
colored currency would make that person appear to be an immoral misanthrope, then
there might be a strong socio-psychological incentive to use it. The psychological
effect might be even stronger when showing the transaction color to the customer
before the transaction is carried out.


# Possible implementation

From a software architecture point of view, there need to be different components
to make the currency system work:

1. A distributed lookup table database to store the colors that can be queried
   during transactions.
2. A collaborative platform, from which the color database can be edited,
   perhaps similar to Github.
3. The cryptocurrency itself (a distributed ledger storing the transactions),
   which is relatively easy to implement, for example as a blockchain in a
   framework like Ethereum.

## The color database or "color blockchain"

The color database gives a unique ID to each product. It is updated frequently
and always returns the current color when queried. It may or may not be a blockchain.
A blockchain would have added benefits like tamper resistance as well as
version control and traceability. If it is a blockchain, then there would be an
"account" for every product and the acount's balance would be the current color
value, which can be changed with "color transactions" by the color determiners,
with added meta-data on why the color is changed. Perhaps, it makes sense to store
not only the resulting color itself, but also "subcolors" for relevant aspects:
social sustainability, environmental sustainability, longevity of the product etc.
Maybe the color database can incorporate user ratings as well, e.g. rating the
usefulness or the quality of the prduct in use.
<!-- XXX Or would the color be part of the cryptocurrency itself, so not an oracle -->
<!-- for a smart contract, but be stored in the transaction blockchain as well? This -->
<!-- way, the currency's blockchain would also provide supply chain tracking, at least -->
<!-- in terms of sustainability. But the blockchain in a traditional sense can only -->
<!-- store transactions that can happen only once, not "products" that can be sold -->
<!-- n times. So, in the blockchain, one should store each transaction's color, -->
<!-- a timestamp, and an ID of the product, but the current product ID's color must -->
<!-- come from an external source. (Or not?) -->
<!-- XXX Or we need two separate blockchains: One for transactions (like bitcoin), one -->
<!-- for colors. In the color blockchain, each product has an "account" and the account's -->
<!-- balance is its color that can change over time. -->

The separate color blockchain would be managed by the individuals determining colors,
the "color miners". Color mining is a task quite different from the transaction
mining, which secures the transaction blockchain. Color mining involves both auditing
or LCA work to gather information backing the color value, but also peer reviewing
other miners' "color transactions" before they become a legitimate part of the
blockchain. Each color miner registers an account on the transaction blockchain, to
which rewards for their work are transferred.

## The collaborative platform

It must be open to anyone. For reading the platform's content, no registration is
necessary. Anyone registering on the platform is becoming a "color miner"
who helps in the determination of colors. Like there can be "projects" in Github,
there are workspaces for each product and service. One has to make sure that there is only
one "genuine" product workspace and that changes to the color of a product are
only done by trusted members. For this purpose, one could introduce a user permission
system like a "web of trust", where only high rated members have permission to
create new work spaces and assign write permissions of certain workspaces to
certain members. The downside is that this makes the system more complicated,
increases the threshold to add new products and introduces a hierarchy. Any better
ideas are appreciated.


## The cryptocurrency or "transaction blockchain"

Ethereum is a programming framework for smart contracts. Smart contracts are a
generalization of cryptocurrencies, where a blockchain can be used for different
purposes. Ethereum is completely independent of the most prominent cryptocurrency,
bitcoin, and has its own blockchain and its own cryptocurrency called *ether*,
which is not used as a currency, but merely as a means of carrying out smart contracts.
With smart contracts, computer code can be attached to the blockchain, which is
executed by the miners when the transaction is validated. Therefore, it can be
made sure that the content of a contract is carried out as intended and this is
stored, secure from manipulation, in the blockchain.

The most simple thing that can be implemented with Ethereum is a new cryptocurrency.
Taking a kind of *Hello World!* example of an Ethereum smart contract [@etherdoc]
and changing it slightly, one arrives at a first rather trivial implementation
of the new cryptocurrency:



# Conclusion

Leveraging the wisdom of crowds and the power of cryptocurrencies, perhaps one can reach a societal consensus towards a world that is more worth living in for everyone, preserving life for future generations, while not giving up modern lifestyles.

These are the requirements that must be fulfilled by the alternative currency:

1. It must feature a significantly more sustainable and/or fair economy.
2. It must be based on societal consensus to be accepted by (almost) everyone.
3. It must be very easy to use.
4. It must be fun to use and/or provide a real advantage.
5. It must be relatively hard to break the rules or abuse the system.

I assume that (1) would be a consequence of all the other points fulfilled, if the currency would be designed similar to what is described above. (2) is the main problem remaining to be solved: how to determine the external costs of goods in a socially acceptable way. The idea to use crowdsourcing was only vaguely described here. More thoughts on this would be greatly appreciated. (5) would only partly be fulfilled by the use of cryptocurrencies, how to secure the crowdsourcing against abuse is another problem to be solved. (3) and (4) have only been merely touched on the surface.

Therefore, we are still at the very beginning, but I hope that something good will evolve out of this.

# References